
As of 2024-04-19 the GapAnalysis R package is not working efficiently due to the dependency on deprecated spatial libraries (sp, raster). 

The package is expected to be updated in at some time in 2024. 

In the mean time, this functions described below can be used to alternative to the GapAnalysis package. 

This will require some working knowledge of github and sourcing functions, making it a generally less user friendly, though more stable application of the Gap Analysis conservation assessment. 

Please direct questions to Dan Carver -- carverd@colostate.edu 



## Downloading the working repository 

Clone or download and extract the following [git repository](https://github.com/dcarver1/cwr_wildgrapes)

The following functions should be sourced into your working environment 

```{r}
### your specific path will differ based on your current working directly and the relative path to the downloaded repository 
# some data processing functions stored in a single file 
source("~/Documents/cwr_wildgrapes/R2/dataProcessing/data_processing_functions.R")
# g_buffer 
source("~/Documents/cwr_wildgrapes/R2/modeling/create_buffers.R")
source("~/Documents/cwr_wildgrapes/R2/modeling/cropG_buffer.R")

# gap analysis function 
## list all options 
f1 <- list.files(
      path = "~/Documents/cwr_wildgrapes/R2/gapAnalysis",
      pattern = "\\.R$", # \\ensure file extension. $ensures ends with. Avoids .Rmd
      full.names = TRUE,
      recursive = TRUE )
# source
for(i in f1){
  print(i)
  source(i)
}
```

## supporting libraries 
Load supporting libraries using pacman 

```{r}
# optional but good 
# install.packages("pacman")
pacman::p_load("terra", "dplyr", "sf","devtools")
```


## testing data 

We will be using the datasets from the gap analysis package for examples. 
If you are running on your own I'd suggesting changing the source here. 

Because of the issues with dependency, GapAnalysis can not be reliably download via CRAN so we will pull and compile from github. 

```{r}
# remotes::install_github("CIAT-DAPA/GapAnalysis")
library(GapAnalysis)
```
Pull in some example data 

The [full ecoregion file](https://geospatial.tnc.org/datasets/b1636d640ede4d6ca8f5e369f2dc368b/about)

```{r}
# pulls the ecoregion and protected areas dataset from a dataverse repository 
GapAnalysis::GetDatasets()

# assign ecoregion 
## note this is a subset of the full file that can be download from the link above  
ecos <- GapAnalysis::ecoregions |>
  sf::st_as_sf()
# assign protected areas 
proAreas <- GapAnalysis::ProtectedAreas |>
  terra::rast()|>
  terra::project(ecos) # reproject the dataset to match ecos --- some to be some issues when moving to terra and maintaining the projection
# change the name to match the requirements of functions 
names(proAreas) <- "layer"
# occurrance records -- store as a dataframe 
occData <- GapAnalysis::CucurbitaData 
# distributions rasters 
rasters <- GapAnalysis::CucurbitaRasters |>
  terra::rast()

```


## Single Species Example 

```{r}
# subset species data
occFilter <- occData |>
  dplyr::filter(species == "Cucurbita_digitata") |>
  dplyr::mutate(databaseSource = "NA") # required for the generateCounts functions 
names(occFilter) <- c("taxon", "latitude", "longitude", "type", "databaseSource")


# subset raster and project to crs of ecos ( this is was is used to define crs of points, but as long as your data have the CRS you should be fine)
raster <- rasters$digitata |> 
  terra::project(ecos)
# reset the name of the raster for some specific indexing within gap analysis functions 
names(raster) <- "Threshold"

## preping some data for the gap analysis functions 
### counts is used for the SRSex measures
counts <- generateCounts(speciesData = occFilter)
### make a spatial representation 
sp1 <-createSF_Objects(speciesData = occFilter)
# optional remove duplicated points will not effect your SRSex score  
sp1 <- removeDuplicates(sp1)
### define natural areas, efficency step so you do have to pass the full ecoregion file to functions. 
natArea <- nat_area_shp(speciesPoints = sp1, ecoregions = ecos)
### or (recommend if you did not limit your model extent by ecoregions. )
# natArea <- ecos 


# generate some measures 
## exsitu
thres <- raster |>
      terra::crop(natArea) |>
      terra::mask(natArea)

### srsex
srsex <- srs_exsitu(sp_counts = counts)
### grsex 
gBuffer <- create_buffers(speciesPoints = sp1,
               natArea = natArea,
               bufferDist = 50000, 
               templateRast = thres)
gCropBuffer <- cropG_Buffer(ga50 = gBuffer, 
                            thres = thres)
grsex <- grs_exsitu(speciesData = sp1,
                    ga50 = gCropBuffer,
                    thres = thres)
### ersex
ersex <- ers_exsitu(speciesData = sp1, 
                    thres = thres,
                    natArea = natArea,
                    ga50 = gCropBuffer
                    )
### fcsex 
fcsex <- fcs_exsitu(srsex = srsex,grsex = grsex,ersex = ersex,noModel = FALSE)


## insitu 
### srsin 
srsin <- srs_insitu(occuranceData = sp1,
                    thres = thres, 
                    protectedArea = proAreas)
### grsin 
grsin <- grs_insitu(occuranceData = sp1,
                    protectedArea = proAreas,
                    thres = thres)
### ersin
ersin <- ers_insitu(occuranceData = sp1,
                    nativeArea = natArea,
                    protectedArea = proAreas,
                    thres = thres)
### fcsin 
fcsin <- fcs_insitu(srsin = srsin,
                    grsin = grsin,
                    ersin = ersin,
                    noModel = FALSE)

### combined models
fcsCombined <- fcs_combine(fcsin = fcsin,
                           fcsex = fcsex)

```



## multiple species 

I'd recommend just tossing the workflow above into a for loop. You'll need to be carefull with the indexing, specifically of the raster elements. 

