---
title: Taxon level modeling, preliminary threat assessment, and conservation gap analysis summary
author: Daniel P. Carver and Colin K. Khoury 
output:
  html_document:
      css: style.css
date: "Report generated on `r Sys.Date()`"
params:
  reportData: NA
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
pacman::p_load("leaflet","data.table", "tidyr", "dplyr","plotly","DT")

### function for control number of character after the decimal place 
sigfig <- function(vec, n=3){ 
### function to round values to N significant digits
# input:   vec       vector of numeric
#          n         integer is the required sigfig  
# output:  outvec    vector of numeric rounded to N sigfig

formatC(signif(vec,digits=n), digits=n,format="fg", flag="#") 

} 



# transform data
# data <- reportData # testing only 
data <- params$reportData
species <- data$occuranceData$taxon[1]
```

# Conservation Summary of `r species`

Plot of the exsitu and insitu conservation scores. [Definitions of Conservation Scores]

```{r echo=FALSE, message=FALSE, warning=FALSE,fig.align='right',out.width="100%"}
d2 <- rbind(data$fcsex, data$fcsin) %>%
  mutate(Category = c("Exsitu", "Insitu"))

d3 <- d2 %>%
  pivot_longer(cols = c("SRS","GRS","ERS","FCS"), names_to = "Metric")
# define the combined score for the species 
d3[9, ] <- list(d3$ID[1], d3$FCS_Score[1],"Combined","FCSc",as.numeric(mean(d3$value[4],d3$value[8])))
#
d3$Metric <- factor(d3$Metric, levels = c("SRS","ERS","GRS","FCS","FCSc"))


plot_ly(d3,
         x = ~Metric,
         y = ~value,
         color = ~Category)%>%
  layout(xaxis = list(title = "", tickangle = 0),
         yaxis = list(title = ""),
         margin = list(b = 100),
         barmode = 'group',
         title = list(text = "<b>Conservation Score Summary</b>",
                      pad = list(b = 20, t = 100 )),
         hovermode = 'compare')

```

*Note: Score values of zero will not appear on the chart.*

# Summary of Occurrence Data

This table summaries the observation data used in the modeling process. [Definition of Occurrence Data Categories]

```{r  echo=FALSE, message=FALSE, warning=FALSE}
# alter Counts data for better visualizations
c1 <- data$countsData %>%
  dplyr::select(Taxon = species,
                'Observations' = totalRecords,
                'Observations with Lat/Long' = totalUseful, 
                'Germplasma Records(G)' = totalGRecords,
                'Herbarium Records(H)' = totalHRecords,
                'Unique Data Sources' = numberOfUniqueSources
                )%>%
  dplyr::mutate(Taxon = stringr::str_replace_all(Taxon, pattern = "_", replacement = " "))


DT::datatable(c1,
              rownames = FALSE,
              class = "compact")

```

# Map of Species Data

Layers can be added to the map using the toggle buttons on the left side. The first layer turned on will be in the bottom position on the map.

You can use this behavior to construct visualizations that convey specific conservation assessments. [Building Conservation Assessment Maps]

```{r echo=FALSE, message=FALSE, warning=FALSE}

# point data
occData <- data$occuranceData %>%
  dplyr::mutate(popup=paste0("<br/><b>Taxon:</b> ", taxon,
                             "<br/><b>Source:</b> ", databaseSource,
                          "<br/><b>Collector Code:</b> ", institutionCode,
                          "<br/><b>Collection Type:</b> ", type)) |>
  mutate(color = case_when(
    type == "H" ~ "#1184D4",
    type == "G" ~ "#6300F0"))
## change this until do that we have a assigned values based on the type 
# pointColor <- colorFactor(palette = c("#1184D4","#6300F0"), domain = occData$type)
gBuf <- data$g_bufferCrop


# predicted presence map
threshold <- data$binaryMap
thresColor <- colorNumeric(palette = c( "#FFFFFF80","#45B320"), 
                           domain = values(threshold),
                           alpha = TRUE,
                           na.color = "transparent")

# protected area
protectedArea <- data$protectedArea
protectColor <- colorNumeric(palette = "#B7B648", domain = values(protectedArea), na.color = "transparent")

#buffered G area
gBuf <- data$g_bufferCrop
if(class(gBuf) != "character"){
  bufColor <- colorNumeric(palette = "#6300F0", domain = values(gBuf),na.color = "transparent")
}

#ecoregions
ecoReg <- data$naturalArea
ecoColor <- colorFactor(palette = "Set1",domain = ecoReg$ECO_ID_U)

#distribution palette 
distColor <- c("#ffffd4","#fed98e","#fe9929","#d95f0e","#993404")

# Mean distribution map 
meanDist <- data$projectedResults$mean
meanPal <- colorNumeric(palette = distColor,domain = values(meanDist), na.color = "transparent")
# Median distribution map 
medianDist <- data$projectedResults$median
medianPal <- colorNumeric(palette = distColor,domain = values(medianDist), na.color = "transparent")

# standdard devation map 
stdevDist <- data$projectedResults$stdev
stdevPal <- colorNumeric(palette = distColor,domain = values(stdevDist), na.color = "transparent")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# leaflet Map
## need to got to documentation for specifics on functions https://github.com/Leaflet/Leaflet.markercluster
m1 <- leaflet(width = "100%")%>%
  addTiles()%>%
  addCircleMarkers(data = occData,
             color = ~color,
             opacity = 1,
             radius = 1,
             group = "Observations",
             stroke = 1,
             popup = ~popup,
             clusterOptions = markerClusterOptions(
               zoomToBoundsOnClick = FALSE,
               disableClusteringAtZoom = 6,
                showCoverageOnHover = TRUE,
                  spiderLegPolylineOptions = list(weight = 1.5, color = "#08FE62", opacity = 0.5)))%>%
  addLegend(position = "bottomleft",
            color = c("#1184D4","#6300F0"),
            labels = c("Herbarium", "Germplasm"),
            title = "Observation Type",
            group = "Observations")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#raster elements
m2 <-m1 %>%
  # threshold 
  addRasterImage(x = raster(threshold), colors = thresColor, group = "Expected Habitat", method = "ngb")%>%
    addLegend(position = "bottomright",
            color = c( "#FFFFFF","#45B320"),
            labels = c("Native Area", "Expected Habitat"),
            title = "Modeled Habitat",
            group = "Expected Habitat")%>%
  #protected area
  addRasterImage(x = raster(protectedArea), colors = protectColor, group = "Protected Area", method = "ngb")%>%
     addLegend(position = "bottomleft",
               color = "#B7B648",
            labels = "Protected Area",
            group = "Protected Area")%>%
  #mean 
  addRasterImage(x = meanDist, 
                 colors = meanPal,
                 group = "Mean Distribution Model",
                 method = "ngb")%>%
    addLegend(position = "bottomright",
            color = distColor,
            title = "Mean Distribution Model",
            labels = c("low","","","","high"),
            group = "Mean Distribution Model")%>%
  #median 
  addRasterImage(x = medianDist, 
                 colors = medianPal,
                 group = "Median Distribution Model",
                 method = "ngb") %>%
    addLegend(position = "bottomright",
            color = distColor,
            title = "Median Distribution Model",
            labels = c("low","","","","high"),
            group = "Median Distribution Model")%>%
    #stdev 
  addRasterImage(x = stdevDist, 
                 colors = stdevPal,
                 group = "SD of Distribution Model",
                 method = "ngb")%>%
    addLegend(position = "bottomright",
            color = distColor,
            title = "SD of Distribution Model",
            labels = c("low","","","","high"),
            group = "SD of Distribution Model")%>%
    # add polygons
      addPolygons(
        data = ecoReg,
        color = ~ecoColor(ECO_ID_U),
        popup = ~ECO_NAME ,
        group = "Ecoregions")


```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# buffer specific formating
if(class(gBuf) != "character"){
  m3 <- m2 %>%  #G buf
  addRasterImage(x = raster(gBuf), colors = bufColor, group = "Buffer Germplasm Collections", method = "ngb")%>%
    addLegend(position = "bottomright",
            color = "#6300F0",
            labels = "Buffer Germplasm Collections",
            group = "Buffer Germplasm Collections")%>%
  addLayersControl(position = "topleft", options = list(collapsed = FALSE,
                                                        autoZIndex = FALSE),
    overlayGroups = c("Expected Habitat","Observations","Protected Area",
                      "Buffer Germplasm Collections", "Ecoregions",
                      "Mean Distribution Model","Median Distribution Model",
                      "SD of Distribution Model")
  )%>%hideGroup( c("Protected Area","Buffer Germplasm Collections", "Ecoregions",
                   "Mean Distribution Model","Median Distribution Model",
                   "SD of Distribution Model"))
}else{
  m3 <- m2 %>%
      addLayersControl(position = "topleft", options = list(collapsed = FALSE,
                                                        autoZIndex = FALSE),
    overlayGroups = c("Expected Habitat","Observations","Protected Area",
                      "Ecoregions",
                      "Mean Distribution Model","Median Distribution Model",
                      "SD of Distribution Model")
  )  %>%hideGroup( c("Protected Area", "Ecoregions",
                   "Mean Distribution Model","Median Distribution Model",
                   "SD of Distribution Model"))

}

# add control groups
m3 



```

### Statistical Summary of the Models.

Median models across MaxEnt replicates were evaluated using the area under the receiver operating characteristic curve (AUC), the SD of the AUC across replicates, and the proportion of the potential distribution model with an SD of the replicates \>0.15

```{r, echo=FALSE, eval=TRUE}
# set the sigfig in the table
eval1 <- data$modelEvaluation %>%
  dplyr::select(
    "AUC" = "AUCtest" ,
    "Normalized AUC" = "nAUC",
    "Calibrated AUC" = "cAUC",
    "Threshold Value" = "threshold_test",
    "Sensitivity"  ="sensi_test",
    "Specificity" ="speci_test", 
    "Mathews Correlation" = "matthews.cor_test",
    "Cohen's kappa" = "kappa_index_test" 
    )%>%
  dplyr::mutate_all(.funs = sigfig)


DT::datatable(eval1,
              class = "hover")
```

<br>

<br>

<br>

<br>

## Definitions of Conservation Scores

### Sampling Representativeness Score (SRS)

**\*Insitu** : The sampling representativeness score in situ (SRSin) calculates the proportion of all occurrences of a taxon within its native range that fall within a protected area.

**Exsitu** : The sampling representativeness score ex situ (SRSex) calculates the ratio of germplasm accessions (G) available in ex situ repositories to reference (H) records for each taxon, making use of all compiled records irrespective of whether they include coordinates.

### Geographic Representativeness Score (GRS)

**Insitu** : The GRSin compares the area (in km2) of the distribution model located within protected areas versus the total area of the model.

**Exsitu** : The GRSex uses 50-km-radius buffers created around each G collection coordinate point to estimate geographic areas already well collected within the distribution models of each taxon and then calculates the proportion of the distribution model covered by these buffers.

### Ecological Representativeness Score (ERS)

**Insitu** : The ERSin calculates the proportion of ecoregions encompassed within the range of the taxon located inside protected areas to the ecoregions encompassed within the total area of the distribution model

**Exsitu** : The ERSex calculates the proportion of terrestrial ecoregions (25) represented within the G buffered areas out of the total number of ecoregions occupied by the distribution model.

### Final Representativeness Score (FCS)

**Insitu** : The FCSin was derived by calculating the average of the three in situ conservation metrics.

**Exsitu** : The FCSex was derived by calculating the average of the three ex situ conservation metrics.

**FCSc-mean** : The FCSc-mean was calculated for each taxon by averaging its final FCSex and FCSin scores. Taxa were then categorized with regard to the two conservation strategies as well as in combination, with UP for further conservation action assigned when FCS \<25, HP assigned when 25 ≤ FCS \< 50, MP when 50 ≤ FCS \< 75, and LP when FCS ≥75.

## Definition of Occurrence Data Categories

**Observations** : The total number of records for the species that were evaluated for used within the study.

**Observations with lat/long**: The total number of records that had a valid latitude and longitude pair and were used for spatial analysis. Non valid latitude longitude pairs included but is not limited to; points not over land masses and records with only one of the two values present.

**Germplasma Records(G)** : Observations in which a living sample (via plant or seed) is present in a preservation focused environment (botanical garden, seed bank). These are observations which could protentially produce viable genetic resource information and are therefore evaluated with the suite of exsit focused conservation metrics.

**Herbarium Records(H)** : Observations that have a supporting herbarium record.

**Unique Data Sources**: The number of unique data base sources from which observations of the species were gathered.

## Building Conservation Assessment Maps

**Geographic representativeness score exsitu** :

1.  Select `Expected Habitat`

2.  Select `Buffer Germplasm Collections`

    The green area outside of the purple is expected habitat that is more then 50km from a germplasm collection. These are potential gaps in the collection of this species.
