---
title: Genus/genepool level visualization of ecogeographic characteristics of taxa
author: Daniel P. Carver, Colin K. Khoury 
output:
  html_document:
      css: style.css
      toc: true
      toc_float: true
date: "Report generated on `r Sys.Date()`"
params:
  inputData: NA
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
# compile all model data for species 
species <- params$inputData$species
data <- params$inputData$data
names <- params$inputData$names
# change to params$inputData$species
```


```{r echo=FALSE, message=FALSE, warning=FALSE}

# generate the plots then append them to a named list
for(i in seq_along(names$shortName)){
  parameter <- names$shortName[i]
  box1 <- generateBoxPlot(data = data,
                names = names,
                parameter = parameter)
  l2 <- list("temp" = box1)
  names(l2) <- parameter

  if(i == 1){
    plots <- l2
    
  }else{
    plots <- c(plots, l2)
  }
}

```

All plots shown below are generated using all available occurrences. The purple points shown on the figures represent the germplasm (G) occurrences for the given taxon. 


## Temperature Related Variables {.tabset}

### Annual mean temperature     

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_01
```

### Mean diurnal temperature range

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_02
```

### Isothermality

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_03
```

### Temperature seasonality (standard deviation)        

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_04
```

### Maximum temperature of warmest month                

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_05
```

### Minimum temperature of coldest month                

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_06
```

### Temperature annual range                            

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_07
```

### Mean temperature of wettest quarter                 

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_08
```

### Mean temperature of driest quarter

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_09
```

### Mean temperature of warmest quarter                 

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_10
```

### Mean temperature of coldest quarter                 

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_11
```

## Precipitation related variables {.tabset}

### Annual precipitation                                

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_12
```

### Precipitation of wettest month                      

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_13
```

### Precipitation of driest month                      

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_14
```

### Precipitation seasonality (coefficient of variation)

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_15
```

### Precipitation of wettest quarter                   

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_16
```

### Precipitation of driest quarter                    

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_17
```

### Precipitation of warmest quarter                   

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_18
```

### Precipitation of coldest quarter                   

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_19
```

## Solar Radiation, Wind, and Topography variables {.tabset}

### srad                                               

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_20
```

### vapr                                               

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_21
```

### wind                                              

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_22
```

### Altitude                                          

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_23
```

### Aspect (North-South)                             

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_24
```

### Aspect (East-West)                                  

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_25
```

### Slope

```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_26
```



## Principal Component Analysis  {.tabset}

```{r echo=FALSE, message=FALSE, warning=FALSE}

library("factoextra")
library("FactoMineR")

# helper function  
visualizeAndSave <- function(pcaplot, title, subtitle, tiltX = FALSE){
  # more direct visualization of the plots 
  p1 <- ggpubr::ggpar(pcaplot,
                title = title,
                subtitle = subtitle,
                legend.title = "Taxon Group",
                legend.position = "top",
                ggtheme = theme_gray()
                )
  p1
  #save result 
  # ggexport(plotlist = p1, 
  #          filename = paste0("outputs/pca/july2024exports/",subtitle,".png"))
           # width = 1200,
           # height = 800,
           # res = 300)
}
```


```{r echo=FALSE, message=FALSE, warning=FALSE}

d2 <- data[,c(30, 3:28)]

attData <- d2 |>
    dplyr::mutate(Taxon = stringr::str_replace_all(string = taxon,
                                                         pattern = "_",
                                                         replace = " ")
  )|>
  dplyr::select(
    Taxon,
    "bio_01","bio_02","bio_03","bio_04","bio_05","bio_06","bio_07","bio_08",
    "bio_09", "bio_10","bio_11","bio_12","bio_13","bio_14","bio_15","bio_16",
    "bio_17","bio_18","bio_19","bio_20", "bio_21","bio_22","bio_23","bio_24",
    "bio_25","bio_26" 
  )
names(attData) <- c("Taxon", names$`Current title`)

# generate the PCA 
pca <- PCA(attData[,-1],scale.unit = TRUE, graph = FALSE)
```

### Eigenvalues 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# eigenvalues/ariance 
eig <- fviz_eig(pca, addlabels = TRUE, ylim = c(0, 50))
visualizeAndSave(pcaplot = eig,
                 title = "Principal Component Analysis- Eigenvalues",
                 subtitle = " ") |> 
    ggplotly(height = 800, width = 1200) 


```

### Correlation Circle

```{r echo=FALSE, message=FALSE, warning=FALSE}


# correlation circle 
var <- get_pca_var(pca)
corCircle <- fviz_pca_var(pca, col.var = "black")
visualizeAndSave(pcaplot = corCircle,
                 title = "Principal Component Analysis - Correlation circle",
                 subtitle = "") |>
  ggplotly(height = 800, width = 1200) 


```

### Quality of Representation 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# quality of representation 
fviz_cos2(pca, choice = "var", axes = 1:2)+
  theme(axis.text.x = element_text(angle = 75))+
  ggtitle("Principal Component Analysis - quality of representation")

```

### Contribution of Variables 

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(gridExtra)
# contributino of variables 
# Contributions of variables to PC1
c1 <- fviz_contrib(pca, choice = "var", axes = 1, top = 10) + 
  theme(axis.text.x = element_text(angle = 75))+
  ggtitle("Contribution of Variables Dim-1")

# Contributions of variables to PC2
c2 <- fviz_contrib(pca, choice = "var", axes = 1, top = 10) + 
  theme(axis.text.x = element_text(angle = 75))+
  ggtitle("Contribution of Variables Dim-2")


grid.arrange(c1, c2, ncol=2)

```

### Ellipse

```{r echo=FALSE, message=FALSE, warning=FALSE}
# pca elispe with groups 
pcaEllipse <- fviz_pca_ind(pca,
             geom.ind = "point", # show points only (nbut not "text")
             col.ind = attData$Taxon, # color by groups
             addEllipses = TRUE, # Concentration ellipses
             mean.point = FALSE,
             legend.title = "Taxon Groups"
)
visualizeAndSave(pcaplot = pcaEllipse,
                 title = "Principal Component Analysis - Ellipse",
                 subtitle = "") |>
  ggplotly(height = 800, width = 1200) 

```


### Confidence Ellipses

```{r echo=FALSE, message=FALSE, warning=FALSE}


# Add confidence ellipses
pcaEllipseConfidence <-fviz_pca_ind(pca, 
             geom.ind = "point", 
             col.ind = attData$Taxon, 
             addEllipses = TRUE, 
             ellipse.type = "confidence",
             mean.point = FALSE,
             legend.title = "Groups"
)
visualizeAndSave(pcaplot = pcaEllipseConfidence,
                 title = "Principal Component Analysis - Confidence Ellipse",
                 subtitle = "Confidence Ellipse")|>
  ggplotly(height = 800, width = 1200) 
```


### Convex Hull

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Convex hull
convexHull <- fviz_pca_ind(pca,
                           geom.ind = "point",
             col.ind = attData$Taxon, # color by groups
             addEllipses = TRUE,
             ellipse.type = "convex",
             legend.title = "Groups"
)
visualizeAndSave(pcaplot = convexHull,
                 title = "Principal Component Analysis - Convex Hull Ellipse",
                 subtitle = "")|>
  ggplotly(height = 800, width = 1200) 
```


### Biplot

```{r echo=FALSE, message=FALSE, warning=FALSE}
# biplot with elipse 
biplot <- fviz_pca_biplot(pca, 
                col.ind = attData$Taxon,
                addEllipses = TRUE, 
                label = "var",
                col.var = "black",
                repel = TRUE,
                legend.title = "Species Group") 

visualizeAndSave(pcaplot = biplot,
                 title = "Principal Component Analysis - Biplot",
                 subtitle = "" )|>
  ggplotly(height = 800, width = 1200) 


```



## Definitions of conservation gap analysis scores

### Sampling Representativeness Score (SRS)

**Ex situ** : The Sampling Representativeness Score *ex situ* (SRS ex situ) calculates the ratio of germplasm accessions (G) available in *ex situ* repositories to reference/voucher (H) records for each taxon, making use of all compiled records irrespective of whether they include coordinates.

**In situ** : The Sampling Representativeness Score *in situ* (SRS in situ) calculates the proportion of all occurrences of a taxon within its native range that fall within a protected area.

### Geographic Representativeness Score (GRS)

**Ex situ** : The Geographic Representativeness Score *ex situ* (GRS ex situ) uses 50-km-radius buffers created around each G collection coordinate point to estimate geographic areas already well collected within the potential distribution models of each taxon and then calculates the proportion of the potential distribution model covered by these buffers.

**In situ** : The Geographic Representativeness Score *in situ* (GRS in situ) compares the area (in km2) of the potential distribution model located within protected areas versus the total area of the model.

### Ecological Representativeness Score (ERS)

**Ex situ** : The Ecological Representativeness Score *ex situ* (ERS ex situ) calculates the proportion of terrestrial ecoregions represented within the G buffered areas out of the total number of ecoregions occupied by the potential distribution model. 

**In situ** : The Ecological Representativeness Score *in situ* (ERS in situ) calculates the proportion of ecoregions encompassed within the range of the taxon located inside protected areas to the ecoregions encompassed within the total area of the potential distribution model. 

### Final Conservation Score (FCS)

**Ex situ** : The Final Conservation Score *ex situ* (FCS ex situ) was derived by calculating the average of the three *ex situ* conservation metrics.

**In situ** : The Final Conservation Score *in situ* (FCS in situ) in was derived by calculating the average of the three *in situ* conservation metrics.

**FCSc-mean** : The Combined Final Conservation Score (FCSc-mean) was calculated for each taxon by averaging its final FCS ex situ and FCS in situ scores. 

Taxa were further categorized with regard to the two conservation strategies as well as in combination, with Urgent Priority (UP) for further conservation action assigned when FCS < 25, High Priority (HP) assigned when 25 ≤ FCS < 50, Medium Priority (MP) when 50 ≤ FCS < 75, and Low Priority (LP) when FCS ≥75.

## Definitions of occurrence data categories

**Occurrences** : The total number of observations or ex situ collections records for the taxon that were evaluated for use within the study.

**Occurrences with lat/long**: The total number of records that had a valid latitude and longitude pair and were used for spatial analysis. Non valid latitude longitude pairs included but were not limited to: points not on land masses, and records with only one of the two values present.

**Germplasm Records (G)** : Occurrences in which a living sample (via plant or seed) is present in an *ex situ* conservation system (i.e., botanical garden, seed bank, genebank, etc.).


**Reference/voucher Records (H)** : Occurrences that have a supporting herbarium or other reference record.

**Unique Data Sources**: The number of unique database sources from which occurrences of the taxon were gathered.
