---
title: Genus/genepool level visualization of species indicator distribution
author: Daniel P. Carver and Colin K. Khoury 
output:
  html_document:
      css: style.css
      toc: true
      toc_float: true
date: "Report generated on `r Sys.Date()`"
params:
  inputData: NA
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
# compile all model data for species 
species <- params$inputData$species
data <- params$inputData$data
names <- params$inputData$names
# change to params$inputData$species

# generate the plots then append them to a named list
for(i in seq_along(names$shortName)){
  parameter <- names$shortName[i]
  box1 <- generateBoxPlot(data = data,
                names = names,
                parameter = parameter)
  l2 <- list("temp" = box1)
  names(l2) <- parameter

  if(i == 1){
    plots <- l2
    
  }else{
    plots <- c(plots, l2)
  }
}

```


### Annual mean temperature     
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_01

```
### Mean diurnal temperature range   
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_02

```
### Isothermality                                     
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_03
```
### Temperature seasonality (standard deviation)        
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_04
```
### Maximum temperature of warmest month                
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_05
```
### Minimum temperature of coldest month                
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_06
```
### Temperature annual range                            
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_07
```
### Mean temperature of wettest quarter                 
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_08
```
### Mean temperature of driest quarter
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_09
```
### Mean temperature of warmest quarter                 
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_10
```
### Mean temperature of coldest quarter                 
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_11
```
### Annual precipitation                                
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_12
```
### Precipitation of wettest month                      
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_13
```
### Precipitation of driest month                      
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_14
```
### Precipitation seasonality (coefficient of variation)
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_15
```
### Precipitation of wettest quarter                   
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_16
```
### Precipitation of driest quarter                    
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_17
```
### Precipitation of warmest quarter                   
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_18
```
### Precipitation of coldest quarter                   
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_19
```
### srad                                               
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_20
```
### vapr                                               
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_21
```
### wind                                              
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_22
```
### Altitude                                          
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_23
```
### Aspect (North-South)                             
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_24
```
### Aspect (East-West)                                  
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_25
```
### Slope
```{r echo=FALSE, message=FALSE, warning=FALSE}
plots$bio_26
```