---
title: Taxon level assessment of occurrence  data
author: Daniel P. Carver, Anne Frances, and Colin Khoury 
output:
  html_document:
      css: style.css
date: "Report generated on `r Sys.Date()`"
params:
  speciesName: NA
  speciesNames: NA
  namedFeatures: NA
  plantsData1: NA
  bonapData: NA
  natureSeverData: NA
  observationData: NA
  countySHP : NA
  stateSHP: NA
  fnaData: NA
---
```{r echo=FALSE, message=FALSE, warning=FALSE}
pacman::p_load("leaflet","data.table", "tidyr", "dplyr","plotly","DT","readr","sf", "tmap",
               "viridis","stringr")
tmap_mode("view")
sf_use_s2(FALSE)
### function for control number of character after the decimal place 
sigfig <- function(vec, n=3){ 
### function to round values to N significant digits
# input:   vec       vector of numeric
#          n         integer is the required sigfig  
# output:  outvec    vector of numeric rounded to N sigfig

formatC(signif(vec,digits=n), digits=n,format="fg", flag="#") 

} 
# for testing 
# speciesName <- "Vitis baileyana"
## troubleshooting 
# speciesName <-"Vitis palmata"
speciesName <- params$speciesName
```


```{r echo=FALSE, message=FALSE, warning=FALSE}

# skip this in troubleshooting 

# reading input datasets --------------------------------------------------
## taxonomic reference 
speciesNames <- params$speciesNames
  # 
  # read_csv(file = "~/Documents/cwr_wildgrapes/data/source_data/taxonomy20230628.csv")
namedFeatures <- params$namedFeatures
  # read_csv(file = "~/Documents/cwr_wildgrapes/data/source_data/nameList.csv")
## county level reference data
plantsData1 <- params$plantsData1
  # read_csv(file ="~/Documents/cwr_wildgrapes/data/source_data/usda_plants/completeVitis.csv")
bonapData <- params$bonapData
  # read_csv("~/Documents/cwr_wildgrapes/data/source_data/bonap.csv")
natureSeverData <- params$natureSeverData
  # read_csv("~/Documents/cwr_wildgrapes/data/processed_occurrence/natureServe.csv")
# valid lat long datasets 
observationData <- params$observationData
  # read_csv("~/Documents/cwr_wildgrapes/data/processed_occurrence/tempDataForCountyMaps_2023118.csv") |> 
  # filter(!is.na(taxon))
fnaData <- params$fnaData

#spatial data 
countySHP <- params$countySHP
# <- read_sf("~/Documents/cwr_wildgrapes/data/geospatial_datasets/counties/ne_10m_admin_2_counties.gpkg")
stateSHP  <- params$stateSHP
# <- read_sf("~/Documents/cwr_wildgrapes/data/geospatial_datasets/states/ne_10m_admin_1_states_provinces.gpkg")|>
#   dplyr::filter(adm0_a3 == "USA")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
speciesList <- dplyr::left_join(x = plantsData1,
                                y = namedFeatures,
                                by = c("plant_symbol" =  "Accepted Symbol"))|>
  dplyr::select( "plant_symbol","Scientific Name")|>
  distinct()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
  species <- speciesList[speciesList$`Scientific Name` == speciesName, ]

  # select county data for species 
  plantsData <- plantsData1 |> 
    filter(plant_symbol %in% species$plant_symbol)|>
    mutate(plantsData = str_sub(geoid, start = 3, end = 7))|>
    select(plantsData)
  bonap <- bonapData |> 
    filter(`Scientific Name` == speciesName)|>
    select(bonap = FIPS)
  natureServe <- natureSeverData[natureSeverData$taxon == speciesName,] |>
    select(natureServe = countyFIPS)

  # combine all the county datasets 
  countyGathered <- countySHP |>
    select("NAME","NAME_ALT","CODE_LOCAL","REGION")|>
    mutate(
      plants = case_when(
        CODE_LOCAL %in% plantsData$plantsData ~ "USDA_Plants"),
      bonap = case_when(
        CODE_LOCAL %in% bonap$bonap ~ "bonap"),
      natureServe = case_when(
        CODE_LOCAL %in% natureServe$natureServe ~ "Nature_Serve"),
      allCountySources = paste0(plants," ",bonap," ",natureServe)
    )|>
  mutate(allCountySources = str_remove_all(allCountySources, pattern = "NA"))|>
    filter(allCountySources != "  ")
  
  # render occ data
  occData <- observationData |>
    dplyr::filter(taxon == speciesName) |>
    dplyr::mutate(type = case_when(
      sampleCategory  == "HUMAN_OBSERVATION" ~ "O",
      TRUE ~ type
    ))

# gather the states listed within FNA
fna1 <- fnaData |> 
  dplyr::filter(`Taxon Name` == speciesName) |>
  dplyr::select(`States from FNA`)|>
  pull()|>
  stringr::str_trim(side = "left")|>
  stringr::str_split(pattern = ",") |>
  unlist()|>
  stringr::str_trim()|>
  str_subset(".+") # remove empty strings 

# need a condition for determining when to run the FNA state filter 
if(length(fna1)== 1 & fna1[1]=="NA"){
  ### no FNA data 
  # use the points and county data to gather the states 
  # select all states from classified data and lat long :FIPS 
  state1 <- paste0("US",unique(occData$stateFIPS)) 
  
  state_CountyData <- stateSHP |>
      filter(iso_3166_2 %in% unique(paste0("US-",countyGathered$REGION)))|>
      select(name)|>
      st_drop_geometry()|> 
      pull()
  
  all_states <- stateSHP |> 
    dplyr::filter(name %in% state_CountyData | code_local %in% state1)
}else{
  # use the fna1 features to select the states 
  state1 <- fna1[fna1 != "NA"]
  # filter the states feature 
 all_states <- stateSHP |>
    filter(name %in% state1)
}

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
  # filter the occurrance data to the states from FNA
  ### 
  occData <- occData |> 
    dplyr::filter(state %in% all_states$name)
  
  
### select states of interset 
# spatial object 
sp1 <- occData |> 
  dplyr::filter(!is.na(latitude))|>
  dplyr::filter(iso3!= "CAN")|>
  dplyr::filter(iso3 != "MEX")|>
  sf::st_as_sf(coords = c("longitude","latitude"), crs = sf::st_crs(stateSHP))

### select all counties of interest 
extratedFIPS <- sf::st_intersection(countySHP, sp1)


countyCounts <- extratedFIPS |> 
  dplyr::group_by(type, CODE_LOCAL)|>
  st_drop_geometry() |> 
  dplyr::summarise(count = n())
  

```





```{r echo=FALSE, message=FALSE, warning=FALSE}
# select all the counties from the states of interest
countyState <- countySHP |> 
  dplyr::filter(REGION %in% all_states$postal)|>
  dplyr::select(NAME, FIPS, REGION)|>
  mutate(
    inCountyData = case_when(
      FIPS %in% paste0("US",countyGathered$CODE_LOCAL) ~ 1,
      TRUE ~ 0
    )
  )
countyState$fips2 <- str_remove_all(string = countyState$FIPS, pattern = "US")

# filter spatial data based on year recorded 
sp2 <- sp1 |> 
  dplyr::filter(is.na(yearRecorded) | yearRecorded >= 1970)
older <- sp1 |> 
  dplyr::filter(yearRecorded < 1970)


for(i in c("O","H","G")){
  c2 <- countyCounts |> 
    dplyr::filter(type == i)
  type <- i
  
  c2 <- c2[,2:3]
  names(c2) <- c("fips2", type)
  countyState <- left_join(countyState, c2, by = "fips2")
  
}

# append the older data count 
df2 <- countyState[unlist(sf::st_intersects(older, countyState)), "FIPS"]|>
    st_drop_geometry()|>
    group_by(FIPS)|>
    dplyr::summarise(pre1970 = n())
# join to county feature 
  countyState <- left_join(x = countyState, y = df2, by = "FIPS")


# join the counties(non point based records) with the spatial obervataions 
c2 <- countyGathered |>
  st_drop_geometry() |>
  select(CODE_LOCAL, 
         `USDA Plants` = plants,
         BONAP = bonap, 
         `Nature Serve` = natureServe)|>
  mutate(`USDA Plants` = 
           case_when(
             !is.na(`USDA Plants`)~ 1),
         BONAP = 
           case_when(
             !is.na(BONAP)~ 1),
         `Nature Serve` = 
           case_when(
             !is.na(`Nature Serve`)~ 1)
           )

countyState <- countyState |>
  left_join(y = c2, by = c("fips2"="CODE_LOCAL"))

# add any record column
countyState <- countyState |>
  rowwise() |>
  mutate(anyRecord = sum(inCountyData,O,H,G,pre1970,`USDA Plants`, 
                         BONAP,`Nature Serve`, na.rm = T))

```

## Map of Species Occurrences for `r speciesName`

The map below summarizes all occurrence data at the country level. This includes 
- Observational data : Validated records from INaturalist
- Herbarium data : Occurrence data with associated herbariumspecimens
- Germplasm data : Occurrence data with living specimens stored in genebanks or botanical gardens 
- USDA Plants : County level reference of species presence
- BONAP : County level reference of species presence
- Nature Serve : county level reference of species presence 

### Spatial Reference
When Latitude and longitude data is present for herbarium and germplasm data, this information is used to spatial derive the county location. When not spatial reference is provide and attempt was made to pull a county reference from attribute data associated with the record. If no certifiable county information was included, the occurrence was drop from the analysis. 

### Flora of North America Mask 
If state level distribution information was provided for a given specific species, that information was used to filter the states in which counties are evalutated. 

### Determining Occupency 

Two maps have been generate to estimate the likelihood of species occurrence within a give county. 

#### Un weighted map 
The goal of this map is to make as little assumptions about the value or impact of a specific record. 

- Observational data : 1 per observation
- Herbarium data : 1 per observation
- Germplasm data : 1 per observation
- USDA Plants : 1 per reference
- BONAP : 1 per reference
- Nature Serve : 1 per reference

For example a county with 3 observational records, 1 germplasm record, and referenced by Nature Serve would have a score of 5. 

#### Weighted Map 
The map was generated using the same input data as the unweighted map, but more consideration was given to the expected significance of a record for a specific data source. In this example any records that were collected prior to 1970 do not contribute to the score.  

- Observational data : 0.25 per observation 
- Herbarium data : 0.5 per observation
- Germplasm data : 1 per observation
- USDA Plants and Bonap combined : 0.5 per reference
- Nature Serve : 1 per reference

The USDA Plants and Bonap measures were combine to a single feature due to the similarity in methods for creating these datasets. If a county was referenced by USDA Plants, Bonap, or both USDA Plants and Bonap it receives a score of 0.5

For example a county with 3 observational records, 1 germplasm record, and referenced by Nature Serve would have a score of 3. 

```{r echo=FALSE, message=FALSE, warning=FALSE}


### county data specifically 

# filter counties based on usda plants data 
  counties <- countyGathered |>
    mutate(popup = paste0("<b>Name:</b> ",NAME_ALT , 
                         "<br/><b>FIPS:</b> ", CODE_LOCAL,
                         "<br/><b>ID Source(s):</b> ", allCountySources))
  
  
# counties caputered by occurrences and county data 
countyClass <- countyState |> 
  dplyr::mutate(
    Ow = O*0.25,
    Hw = H*0.5,
    Cw = case_when(
      `USDA Plants` == 1 ~ 0.5,
      `BONAP` == 1 ~ 0.5,
      TRUE ~ 0 
    )
  )|>
  rowwise() |>
  mutate(weightedSum = sum(Ow,Hw,G,Cw,`Nature Serve`, na.rm = T))|>
  dplyr::select(-c(Ow,Hw,Cw))|>
  dplyr::mutate(
    popup = paste0(
        "<b>County Name: </b>", NAME,
        "<br/> <b>State: </b>", REGION,
        "<br/> <b> Listed in County Source Data: </b>", inCountyData,
        "<br/> <b> Listed in USDA Plants: </b>", `USDA Plants`,
        "<br/> <b> Listed in BONAP: </b>", `BONAP`,
        "<br/> <b> Listed in Nature Serve: </b>", `Nature Serve`,
        "<br/> <b> Number of Observations Only Records: </b>", O,
        "<br/> <b> Number of herbarium Records: </b>", H,
        "<br/> <b> Number of germplasm  Records: </b>", G,
        "<br/> <b> Number of Records before 1970: </b>",pre1970,
        "<br/> <b> All Records: </b>",anyRecord,
        "<br/> <b> Weighted Records: </b>",weightedSum 
      )
  )|>
  dplyr::mutate(
    weightedSum = case_when(weightedSum == 0 ~ NA,
                            TRUE ~ weightedSum),
    anyRecord = case_when(anyRecord == 0 ~NA,
                          TRUE ~ anyRecord)
  )

## color palatte 
### Good set to use for the visualization  
brbg <- c("#a6611a","#dfc27d","#f5f5f5",'#c7eae5','#80cdc1','#35978f','#01665e')
# Specific elements used in the reclass
brbgWeighted <- brbg[c(1,4,5,6,7)]
weightVals <- c("<1","1-2","2-3","3-4",">=4")
brbgAll <- brbg[c(1,2,4,5,6,7)]
allVals <- c("1", "2-3","3-5","5-8","8-11",">12")
# trying to assign classes directly 
countyClass <- countyClass |>
  dplyr::mutate(
    weightedColor= case_when(
      is.na(weightedSum) ~ "#d9dbde",
      weightedSum < 1 ~ brbg[1],
      weightedSum >= 1 & weightedSum < 2 ~ brbg[4],
      weightedSum >= 2 & weightedSum < 3 ~ brbg[5],
      weightedSum >= 3 & weightedSum < 4 ~ brbg[6],
      weightedSum >= 4  ~ brbg[7]
    ),
    allColor = case_when(
      is.na(anyRecord) ~ "#d9dbde",
      anyRecord == 1 ~ brbg[1],
      anyRecord >= 2 & anyRecord < 3 ~ brbg[2],
      anyRecord >= 3 & anyRecord < 5 ~ brbg[4],
      anyRecord >= 5 & anyRecord < 8 ~ brbg[5],
      anyRecord >= 8 & anyRecord < 12 ~ brbg[6],
      anyRecord >= 12  ~ brbg[7]
    )
  )

# Create a continuous palette function
palAll <- colorNumeric(
  palette = "BrBG",
  domain = countyClass$anyRecord,
  na.color = "#d9dbde")
palWeighted <- colorNumeric(
  palette = "BrBG",
  domain = countyClass$weightedSum,
  na.color = "#d9dbde")


points <- sp1 |>
  dplyr::mutate(
     popup =paste0(
    "<b>Taxon: </b>", taxon,
    "<br/> <b>Data Source: </b>", databaseSource,
    "<br/> <b> Record ID: </b>", sourceUniqueID,
    "<br/> <b> Type: </b>", type,
    "<br/> <b> Year: </b>", yearRecorded,
    "<br/> <b> Location Description: </b>", localityInformation,
    "<br/> <b> External Reference: </b>",collectionSource ),
    color = case_when(
      type == "O" ~ "#fc8d62",
      type == "H" ~ "#8da0cd",
      type == "G" ~ "#66c2a5"
    ))


# Map 
## need conditions for the various types inputs 
m1 <- leaflet(width = "100%",
              sizingPolicy = leafletSizingPolicy(defaultWidth = "100%",
                                                 defaultHeight = 2000,
                                                 padding = 10),
              options = leafletOptions(minZoom = 3))|>
    addTiles(urlTemplate = 'https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/World_Light_Gray_Base/MapServer/tile/{z}/{y}/{x}', 
             attribution =  'Tiles &copy; Esri &mdash; Esri, DeLorme, NAVTEQ'
               )|>
    addMapPane("allRecords", zIndex = 415) |>
    addMapPane("weightedRecords", zIndex = 418) |>
    addMapPane("points", zIndex = 420)

# add spatial features to map
  m2 <- m1 |>
    addPolygons(data = countyClass,
                color = "white",
                fillColor =~allColor,
                popup = ~popup,
                weight = 0.8,
                fillOpacity = 0.6,
                opacity = 0.8,
                group = "All Records",
                options = pathOptions(pane = "allRecords"))|>
    addPolygons(data = countyClass,
                color = "white",
                fillColor = ~weightedColor,
                popup = ~popup,
                weight = 0.8,
                fillOpacity = 0.6,
                opacity = 0.8,
                group = "Weighted Records",
                options = pathOptions(pane = "weightedRecords"))|> 
    addCircleMarkers(data =  points |> filter(type =="O"),
             color = ~color,
             opacity = 1,
             radius = 2,
             stroke = 1,
             popup = ~popup,
             group = "Observation",
             options = pathOptions(pane = "points"))|>
    addCircleMarkers(data = points |> filter(type =="H"),
             color = ~color,
             opacity = 1,
             radius = 2,
             stroke = 1,
             popup = ~popup,
             group = "Herbarium",
             options = pathOptions(pane = "points")
               )|>
    addCircleMarkers(data = points[points$type =="G",],
             color = ~color,
             opacity = 1,
             radius = 2,
             stroke = 1,
             popup = ~popup,
             group = "Germplasm",
             options = pathOptions(pane = "points")
               )|>
    addLegend(color = brbgAll,
              labels = allVals, # countyClass$anyRecord[!is.na(countyClass$anyRecord)],
              group = "All Records",
              title = "All Records Counted")|>
    addLegend(colors = brbgWeighted,
              labels = weightVals,
              # values = ~weightedSum, # countyClass$weightedSum[!is.na(countyClass$weightedSum)],
              na.label = NA,
              group = "Weighted Records",
              title = "Weighted Records")|>
  addLayersControl(
      overlayGroups = c("All Records","Weighted Records","Observation", "Herbarium", "Germplasm"),
      options = layersControlOptions(collapsed = FALSE)
    )|>
  hideGroup(c("Weighted Records","Observation","Herbarium","Germplasm"))


m2
```




<!-- Old section 
## Map of Species Occurrences for `r speciesName`

The map below summarizes all occurrence data at the county level. All counties within a state where are observation was described have been included. It is possible that some of the counties, even though they may not have known occurrences, should be considered habitat for a given taxon. 
Selecting a county will provide the name and summary of the observations within. 

### County Classifications Rules 

> Blue: color counties are of low priority for review for one of the following reasons: 
<br>
- germplasm  sample has been collected from the county after 1970 
<br>
- Two herbarium samples have been collected from the county after 1970
<br>
- The county was identified by USDA Plants or Bonap and contains at least one herbarium record
<br>
- The county was indentified by Nature Serve 

<br>

> Orange: color counties are of high priority for review for one of the following reasons. 
<br>
- All observations were made before 1970
<br>
- A single herbarium record was noted after 1970
<br>
- Only observational data has been noted within the county. 
<br>
- Only defined by USDA Plants

> Grey:  color counties are locations where no observations have been noted but an observation has been recorded in the state. This is mostly here so individuals can identify counties worth of inclusion despite the lack of observational data. 


### Classification of Observational Data

> Orange: color points represent observational data, such as iNaturalist reports. 

<br>

> Blue: colored points represent herbarium data

<br>

> Green: colored points represents gremplasm sample. Most species do not have any germplasm   records.  

<br>


```{r echo=FALSE, message=FALSE, warning=FALSE}


### county data specifically 

# defined palette for county classificatino 
palDF <- data.frame(ref = c(" bonap ",
                            "USDA_Plants  ",
                            "  Nature_Serve",
                            "USDA_Plants bonap ",
                            "USDA_Plants Nature_serve ",
                            " bonap Nature_Serve",
                            "USDA_Plants bonap Nature_Serve"),
                    pal = c("#1b9e77",
                            "#d95f02",
                            "#7570b3",
                            "#e7298a",
                            "#66a61e",
                            "#e6ab02",
                            "#a6761d")
)
# c(" bonap ","USDA_Plants","Nature_Serve","USDA_Plants bonap ","USDA_Plants Nature_serve "," bonap Nature_Serve","USDA_Plants bonap Nature_Serve")
# c("#1b9e77","#d95f02", "#7570b3", "#e7298a","#66a61e", "#e6ab02", "#a6761d")

# filter counties based on usda plants data 
  counties <- countyGathered |>
    mutate(popup = paste0("<b>Name:</b> ",NAME_ALT , 
                         "<br/><b>FIPS:</b> ", CODE_LOCAL,
                         "<br/><b>ID Source(s):</b> ", allCountySources))|>
  mutate(color = case_when(
    allCountySources == " bonap " ~ "#1b9e77",
    allCountySources == "USDA_Plants  " ~ "#d95f02",
    allCountySources == "  Nature_Serve" ~ "#7570b3",
    allCountySources == "USDA_Plants bonap " ~ "#e7298a",
    allCountySources == "USDA_Plants Nature_serve  " ~ "#66a61e",
    allCountySources == " bonap Nature_Serve" ~ "#e6ab02",
    allCountySources == "USDA_Plants bonap Nature_Serve" ~ "#a6761d"
  ))

palDF_filtered <- palDF |>
  filter(ref %in% unique(counties$allCountySources))
  
  
# counties caputered by occurrences and county data 
countyClass <- countyState |> 
  dplyr::mutate(reviewCounty = case_when(
    G > 0 ~ FALSE, 
    H >= 2 ~ FALSE,
    `USDA Plants` ==1 & H ==1 ~ FALSE,
    BONAP == 1 & H ==1 ~ FALSE,
    `Nature Serve` ==1 ~ FALSE,
    inCountyData == 1 & H ==1 ~ FALSE,
    anyRecord == 0 ~ FALSE, 
    TRUE ~ TRUE
  ), color = case_when(
      anyRecord >0 & reviewCounty  == TRUE  ~ "#fdae61",
      anyRecord >0 & reviewCounty == FALSE ~ "#74add1",
      anyRecord == 0 ~ "#bdbdbd"
    ),
  popup = paste0(
      "<b>County Name: </b>", NAME,
      "<br/> <b>State: </b>", REGION,
      "<br/> <b> Listed in County Source Data: </b>", inCountyData,
      "<br/> <b> Listed in USDA Plants: </b>", `USDA Plants`,
      "<br/> <b> Listed in BONAP: </b>", `BONAP`,
      "<br/> <b> Listed in Nature Serve: </b>", `Nature Serve`,
      "<br/> <b> Number of Observations Only Records: </b>", O,
      "<br/> <b> Number of herbarium Records: </b>", H,
      "<br/> <b> Number of germplasm  Records: </b>", G,
      "<br/> <b> Number of Records before 1970: </b>",pre1970  
    )
)

points <- sp1 |>
  dplyr::mutate(
     popup =paste0(
    "<b>Taxon: </b>", taxon,
    "<br/> <b>Data Source: </b>", databaseSource,
    "<br/> <b> Record ID: </b>", sourceUniqueID,
    "<br/> <b> Type: </b>", type,
    "<br/> <b> Year: </b>", yearRecorded,
    "<br/> <b> Location Description: </b>", localityInformation,
    "<br/> <b> External Reference: </b>",collectionSource ),
    color = case_when(
      type == "O" ~ "#fc8d62",
      type == "H" ~ "#8da0cd",
      type == "G" ~ "#66c2a5"
    ))


# Map 
## need conditions for the various types inputs 
m1 <- leaflet(width = "100%",
              sizingPolicy = leafletSizingPolicy(defaultWidth = "100%",
                                                 defaultHeight = 2000,
                                                 padding = 10),
              options = leafletOptions(minZoom = 3))|>
    addTiles(urlTemplate = 'https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/World_Light_Gray_Base/MapServer/tile/{z}/{y}/{x}', 
             attribution =  'Tiles &copy; Esri &mdash; Esri, DeLorme, NAVTEQ'
               )|>
    addMapPane("countiesCheck", zIndex = 415) |>
    addMapPane("countiesDesc", zIndex = 418) |>
    addMapPane("points", zIndex = 420)

if(nrow(counties) == 0){
  # no county desciption features 
  m2 <- m1 |>
    addPolygons(data = countyClass,
                color = "white",
                fillColor =~color,
                popup = ~popup,
                weight = 0.8,
                fillOpacity = 0.6,
                opacity = 0.8,
                group = "Counties to review",
                options = pathOptions(pane = "countiesCheck"))|>
    addCircleMarkers(data =  points |> filter(type =="O"),
             color = ~color,
             opacity = 1,
             radius = 2,
             stroke = 1,
             popup = ~popup,
             group = "Observation",
             options = pathOptions(pane = "points"))|>
    addCircleMarkers(data = points |> filter(type =="H"),
             color = ~color,
             opacity = 1,
             radius = 2,
             stroke = 1,
             popup = ~popup,
             group = "Herbarium",
             options = pathOptions(pane = "points")
               )|>
    addCircleMarkers(data = points[points$type =="G",],
             color = ~color,
             opacity = 1,
             radius = 2,
             stroke = 1,
             popup = ~popup,
             group = "Germplasm ",
             options = pathOptions(pane = "points")
               )|>
    addLegend(colors = c("#74add1","#fdae61","#bdbdbd"),
              labels = c("Acceptable County","Review County", "Unknown"),
              group = "Counties to review",
              position = "bottomright") |>
    # addLegend(position = "bottomright",
    #           colors = palDF_filtered$pal,
    #           labels = palDF_filtered$ref,
    #           group = "County_Reference",
    #           title = "County Reference Data")|>
  addLayersControl(
      overlayGroups = c("Counties to review","Observation", "Herbarium", "Germplasm "),
      options = layersControlOptions(collapsed = FALSE)
    )|>
  hideGroup(c("Observation","Herbarium","Germplasm "))
}else{
  m2 <- m1 |>
   addPolygons(data = counties,
                color = ~color,
                group = "Counties described",
                label = ~counties$NAME_ALT,
                opacity = 0.2,
                fillOpacity = 1,
                popup = ~popup,
                options = pathOptions(pane = "countiesDesc")) |>
    addPolygons(data = countyClass,
                color = "white",
                fillColor =~color,
                popup = ~popup,
                weight = 0.8,
                fillOpacity = 0.6,
                opacity = 0.8,
                group = "Counties to review",
                options = pathOptions(pane = "countiesCheck"))|>
    addCircleMarkers(data =  points |> filter(type =="O"),
             color = ~color,
             opacity = 1,
             radius = 2,
             stroke = 1,
             popup = ~popup,
             group = "Observation",
             options = pathOptions(pane = "points"))|>
    addCircleMarkers(data = points |> filter(type =="H"),
             color = ~color,
             opacity = 1,
             radius = 2,
             stroke = 1,
             popup = ~popup,
             group = "Herbarium",
             options = pathOptions(pane = "points")
               )|>
    addCircleMarkers(data = points[points$type =="G",],
             color = ~color,
             opacity = 1,
             radius = 2,
             stroke = 1,
             popup = ~popup,
             group = "Germplasm",
             options = pathOptions(pane = "points")
               )|>
    addLegend(colors = c("#74add1","#fdae61","#bdbdbd"),
              labels = c("Acceptable County","Review County", "Unknown"),
              group = "Counties to review",
              position = "bottomright",
              title = "Counties to review") |>
    addLegend(position = "bottomright",
              colors = palDF_filtered$pal,
              labels = palDF_filtered$ref,
              group = "Counties described",
              title = "Counties with Described Occurrence")|>
  addLayersControl(
      overlayGroups = c("Counties described","Counties to review","Observation", "Herbarium", "Germplasm"),
      options = layersControlOptions(collapsed = FALSE)
    )|>
  hideGroup(c("Counties described", "Observation","Herbarium","Germplasm"))
}


m2
```

-->


## Summary of Occurrance Data 
This table includes all records not just those with valid latitude and longitude information. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
occFiltered <- occData %>%
  dplyr::select(
    "taxon",
    "state",
    "county",
    "databaseSource",
    "collectionSource",
    "type",
    "localityInformation",
    "yearRecorded",
     "latitude",
    "longitude")


DT::datatable(occFiltered,
              rownames = FALSE,
              width = "100%",
              filter = c("top")
              )

```

## Summary Table of County Data 

The table below display additional attribute data for all the counties with know observation on the the selected taxon. All columns can be filtered and sorted. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# 
df <- countyClass |>
  st_drop_geometry()|>
  dplyr::filter(anyRecord !=0)|>
  dplyr::select(-color, -popup, -anyRecord)|>
  dplyr::mutate(inCountyData = case_when(
    inCountyData == 1 ~ TRUE,
    TRUE ~ FALSE
  ))|>
  dplyr::arrange(REGION)|>
  rename(
    "County Name" = NAME,
    "GEOID" = "FIPS", 
    "State Abbrevation" = "REGION",
    "County Reference" = "inCountyData",
    "Obersevation Only Records" = O,
    "Heberium Records" = H,
    "germplasm  Records" = G,
    "Pre 1970 Records" = pre1970,
    "USDA Plants" = `USDA Plants`,
    "BONAP" = BONAP,
    "Nature Serve" = `Nature Serve`,
    "Recommend Review" = reviewCounty) 

DT::datatable(df,
              rownames = FALSE,
              filter = c("top")
              )

```

