---
title: Taxon level buffered area and conservation gap analysis summary
author: Daniel P. Carver, Colin K. Khoury, Sarah L. Gora
output:
  html_document:
      css: style.css
date: "Report generated on `r Sys.Date()`"
params:
  reportData: NA
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
pacman::p_load("leaflet","data.table", "tidyr", "dplyr","plotly","DT","RColorBrewer","htmltools")

### function for control number of character after the decimal place 
sigfig <- function(vec, n=3){ 
### function to round values to N significant digits
# input:   vec       vector of numeric
#          n         integer is the required sigfig  
# output:  outvec    vector of numeric rounded to N sigfig

formatC(signif(vec,digits=n), digits=n,format="fg", flag="#") 

}  
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# transform data

testing <- FALSE
# testing <- TRUE
if(testing == TRUE){
  data <- reportData # testing only
  species <- data$occuranceData$taxon[1] # testing
}else{
  data <- params$reportData
  species <- data$occuranceData$taxon[1]
}

printSpecies <- stringr::str_replace_all(string = species, pattern = "_", replacement = " ")

# --- Reduce polygon complexity and coordinate precision to reduce HTML size ---
round_sf_coords <- function(sf_obj, digits = 3) {
  if (is.null(sf_obj) || nrow(sf_obj) == 0) return(sf_obj)
  geom <- sf::st_geometry(sf_obj)
  geom <- sf::st_sfc(lapply(geom, function(x) {
    if (inherits(x, "POLYGON")) {
      sf::st_polygon(lapply(x, function(y) round(y, digits)))
    } else if (inherits(x, "MULTIPOLYGON")) {
      sf::st_multipolygon(lapply(x, function(y) lapply(y, function(z) round(z, digits))))
    } else if (inherits(x, "POINT")) {
      sf::st_point(round(as.numeric(x), digits))
    } else {
      x
    }
  }), crs = sf::st_crs(sf_obj))
  sf::st_geometry(sf_obj) <- geom
  return(sf_obj)
}

if (!is.null(data$naturalArea)) {
  # Simplify polygons predictably by turning off S2 temporarily 
  # (so dTolerance is in degrees: 0.05 deg is ~5.5km)
  s2_state <- sf::sf_use_s2()
  sf::sf_use_s2(FALSE)
  
  is_longlat <- sf::st_is_longlat(data$naturalArea)
  tol <- if (!is.na(is_longlat) && is_longlat) 0.05 else 5000
  data$naturalArea <- sf::st_simplify(data$naturalArea, preserveTopology = FALSE, dTolerance = tol)
  
  sf::sf_use_s2(s2_state)
  
  # Round coordinates (2 digits is ~1.1km precision)
  digits_to_keep <- if (!is.na(is_longlat) && is_longlat) 2 else 0
  data$naturalArea <- round_sf_coords(data$naturalArea, digits = digits_to_keep)
}

if (!is.null(data$occuranceData)) {
  if (inherits(data$occuranceData, "sf")) {
    is_longlat <- sf::st_is_longlat(data$occuranceData)
    digits_to_keep <- if (!is.na(is_longlat) && is_longlat) 3 else 0
    data$occuranceData <- round_sf_coords(data$occuranceData, digits = digits_to_keep)
  }
  if ("longitude" %in% names(data$occuranceData)) {
    data$occuranceData$longitude <- round(data$occuranceData$longitude, 3)
    data$occuranceData$latitude <- round(data$occuranceData$latitude, 3)
  }
}

# remove the g point from MBG for bloodworthiana 
if(species == "Vitis bloodworthiana"){
  data$occuranceData <- data$occuranceData[data$occuranceData$longitude != -100.00000, ]
}


```

# `r paste0('*', printSpecies, '*')`

<br>

## Occurrence records    

The table below shows a summary of the occurrence records for the taxon. A record “with Lat/Long” is one that has a complete set of latitude and longitude values associated with it, which could be used for distribution modeling and the conservation gap analysis. In preparation for the conservation gap analyses, we classified each record based on whether it was an existing *ex situ* germplasm sample from a genebank, botanic garden, or other repository (labeled G, as most records were from genebanks), or a reference observation/voucher specimen (labeled H, as most of these records were from herbaria).

```{r  echo=FALSE, message=FALSE, warning=FALSE}
# alter Counts data for better visualizations
c1_a <- data$countsData |>
  dplyr::select(
    Taxon = species,
    'Occurrences' = totalRecords,
    'Occurrences with Lat/Long' = totalUseful,
    'Germplasm Records(G)' = totalGRecords,
    'Germplasm Records(G) with Lat/Long' = totalGUseful,
    'Reference Records(H)' = totalHRecords,
    'Reference Records(H) with Lat/Long' = totalHUseful,
    'Unique Data Sources' = numberOfUniqueSources
  ) 
# add in the model data counts if they exist (they may be NA for buffer models)
if(is.data.frame(data$modelDataCounts) || is.list(data$modelDataCounts)){
  c2 <- data$modelDataCounts |>
    dplyr::select(
      Taxon = species,
      'Presence Occurrences' = presenceRecords,
      'Background Occurrences' = backgroudRecords,
      'Total Occurrences Used for Buffer' = totalRecords
    ) 
  # join 
  cj <- dplyr::left_join(c1_a, c2, by = "Taxon") |>
    dplyr::mutate(Taxon = stringr::str_replace_all(Taxon, pattern = "_", replacement = " "))  
} else {
  # if no model counts, just use the main counts data and maybe add NA columns so the table structure remains consistent
  cj <- c1_a |>
    dplyr::mutate(
      Taxon = stringr::str_replace_all(Taxon, pattern = "_", replacement = " "),
      'Presence Occurrences' = NA,
      'Background Occurrences' = NA,
      'Total Occurrences Used for Buffer' = NA
    )
}

DT::datatable(cj,
              rownames = FALSE,
              class = "compact",
              options = list(dom = 't'))

```


```{r, echo=FALSE,  message=FALSE, warning=FALSE}
# # resample the raster datasets to 5km for visualization
# ## threshold maps
# thresAgg <- matrix(c(
#   -Inf, 0.5, 0,  # All values from -Infinity up to (but not including) 0.5 become 0
#   0.5,  Inf, 1   # All values from 0.5 (including 0.5) up to +Infinity become 1
# ), ncol = 3, byrow = TRUE)
#
# resampleFactor <- 5
#
# resampleRast <- function(rast, resampleFactor = 5){
#   r1 <- rast |>
#     terra::aggregate(fact = resampleFactor, fun = "mean")  |>
#     terra::classify(thresAgg, include.lowest = TRUE)
#   return(r1)
# }
```

<br>


```{r echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
# Generate all the map data
# point data

occData <- data$occuranceData |>
  dplyr::mutate(popup = paste0("<b>", taxon, "</b><br/>", type, " (", databaseSource, ")")) |>
  mutate(color = case_when(type == "H" ~ "#1184D4", type == "G" ~ "#6300F0"))

webMecOccData <- vect(st_transform(occData, crs = "epsg:3857")) # need to change project

if (data$NoModel == FALSE) {
  # buffered area map
  threshold <- data$binaryMap
  
  # --- Reduce raster resolution for widespread species ---
  if (species %in% c(
    "Vitis riparia",
    "Vitis rotundifolia",
    "Vitis tiliifolia",
    "Vitis_riparia",
    "Vitis_rotundifolia",
    "Vitis_tiliifolia"
  )) {
    resampleFactor <- 5
    
    threshold <- terra::aggregate(threshold,
                                  fact = resampleFactor,
                                  fun = "max",
                                  na.rm = TRUE)
    
    if (!is.null(data$protectedArea)) {
      data$protectedArea <- terra::aggregate(
        data$protectedArea,
        fact = resampleFactor,
        fun = "max",
        na.rm = TRUE
      )
    }
    
    if (!is.null(data$projectedResults$mean)) {
      data$projectedResults$mean <- terra::aggregate(
        data$projectedResults$mean,
        fact = resampleFactor,
        fun = "mean",
        na.rm = TRUE
      )
    }
    if (!is.null(data$projectedResults$median)) {
      data$projectedResults$median <- terra::aggregate(
        data$projectedResults$median,
        fact = resampleFactor,
        fun = "mean",
        na.rm = TRUE
      )
    }
    if (!is.null(data$projectedResults$stdev)) {
      data$projectedResults$stdev <- terra::aggregate(
        data$projectedResults$stdev,
        fact = resampleFactor,
        fun = "mean",
        na.rm = TRUE
      )
    }
    
    if (!is.null(data$g_bufferCrop) &&
        inherits(data$g_bufferCrop, "SpatRaster")) {
      data$g_bufferCrop <- terra::aggregate(
        data$g_bufferCrop,
        fact = resampleFactor,
        fun = "max",
        na.rm = TRUE
      )
    }
  }
  # -------------------------------------------------------
  
  thresColor <- colorNumeric(
    palette = c("#cfcccc70", "#45B320"),
    domain = values(threshold),
    alpha = TRUE,
    na.color = "transparent"
  )
  # generate a threshold mask
  mask1 <- ifel(test = threshold == 1,
                yes = 1,
                no = NA)
  # protected area
  protectedArea <- data$protectedArea
  
  protectColor <-
    colorNumeric(
      palette = "#B7B648",
      domain = values(protectedArea),
      alpha = TRUE,
      na.color = "transparent"
    )
  # protected areas mask
  proMask <- ifel(test = protectedArea == 1,
                  yes = 1,
                  no = NA)
  # insitu map elements
  ## SRSIN
  ## total number of observations per cell of the 50 km buffer area masked against the protect areas layer
  # filter points outside of protected areas
  webMecOccData <- vect(st_transform(occData, crs = "epsg:3857")) # need to change project
  # for the spatial operations with the layer. occData can't have the proj
  # get all the points that are outside of a protected area
  
  
  inThres <- terra::extract(x = mask1, y = webMecOccData) |>
    dplyr::filter(Threshold == 1)
  inPro <- terra::extract(x = protectedArea, y = webMecOccData) |>
    dplyr::filter(is.na(layer))
  
  srs1 <- terra::rasterize(webMecOccData,
                           threshold,
                           fun = "sum",
                           background = 0) |>
    mask(mask1)
  
  totalObsinThres <- nrow(inThres)
  ## total number of observations in protected areas
  totalObsinPro <- nrow(inThres) - nrow(inPro)
  
  ## GRSIN
  ## reclassy protected area
  proReclass <- ifel(test = is.na(protectedArea),
                     yes = 0,
                     no = 1)
  ## add values
  grs1 <- threshold + proReclass
  exPal <- c("#cfcccc80", "#45b320", "#7570b3")
  # for some reason passing the vector exPal to the palette arg does not maintain the alpha reference?
  grsinColor <- colorNumeric(
    palette = c("#cfcccc80", "#45b320", "#7570b3"),
    # purple is a little lighter so the point locations show up clearly
    domain = values(grs1),
    alpha = TRUE,
    na.color = "transparent"
  )
  ## ERSin
  # used on the ERS visualization
  allEcos <- data$naturalArea |>
    dplyr::select("ECO_ID_U", "ECO_NAME")
  # reproject for spatial operations
  webMecAllEcos <- st_transform(allEcos, crs = "epsg:3857")
  ## showcase ecoregions within the 50 km buffer area that do no have a protected area
  ers1 <- terra::zonal(proReclass,
                       vect(webMecAllEcos),
                       fun = "sum",
                       na.rm = TRUE)
  ers1$ECO_ID_U <- allEcos$ECO_ID_U
  ers1$ECO_NAME <- allEcos$ECO_NAME
  
  ## zonal Stats to determine overlap between the ecoregions and the 50 km buffer area
  v1 <- terra::zonal(
    x = threshold,
    z = vect(webMecAllEcos),
    fun = "sum",
    na.rm = TRUE
  )# , exact = TRUE)
  
  ## reassign the eco id record
  v1$ECO_ID_U <- webMecAllEcos$ECO_ID_U
  # get a list of ecoregions within the 50 km buffer area
  ecosInThres <- v1 |>
    dplyr::filter(Threshold > 0) |>
    dplyr::select(ECO_ID_U) |>
    pull()
  
  ## ID ecoregions in 50 km buffer area with no protected areas
  ers2 <- ers1 |>
    dplyr::filter(ECO_ID_U %in% ecosInThres) |>
    dplyr::filter(layer == 0)
  
  if (nrow(ers2) > 0) {
    ## use this list to generate a raster object masked to the
    ersinMap <- webMecAllEcos |>
      dplyr::filter(ECO_ID_U %in% ers2$ECO_ID_U) |>
      terra::rasterize(y = threshold, "ECO_ID_U") |>
      mask(mask = mask1)
    # generate palatte object for map
    ecoPal2 <-  rep(RColorBrewer::brewer.pal(n = 12, name = "Spectral"), 6)
    ecoPal2 <- ecoPal2[1:length(ers2$ECO_ID_U)]
    # need to order the names so they appear as expected
    ers2$color <- ecoPal2
    ers2 <- ers2 |> arrange(desc(ECO_NAME))
    ers2$color <- ecoPal2
    
    ers2$ECO_ID_U <- factor(ers2$ECO_ID_U ,
                            ordered = TRUE,
                            levels = c(ers2$ECO_ID_U))
    
    
    # define a palette option so that we can set a transparent NA value
    ecoInColor <- colorFactor(
      palette = ers2$color,
      domain = ers2$ECO_ID_U,
      na.color =  "#FF000000"
    )
    
    
  } else{
    ersinMap <- "No values"
  }
  
  
  
  # exsitu map elements
  
  #buffered G area
  gBuf <- data$g_bufferCrop
  
  if (class(gBuf) != "character" & data$fcsex$GRS != 0) {
    # resample
    gBuf <- gBuf
    
    bufColor <-
      colorNumeric(
        palette = "#6300F0",
        domain = values(gBuf),
        na.color = "transparent"
      )
    # GRSex map element
    m1 <- cbind(NaN, 0)
    gbuf2 <- terra::classify(x = gBuf, m1)
    grsexMap <- threshold + gbuf2
    # need to defined this outside of the palette function so it can be called again in the legend
    grsexColor <- colorNumeric(
      palette = c("#cfcccc80", "#45b320", "#7570b3"),
      # purple is a little lighter so the point locations show up clearly
      domain = values(grsexMap),
      alpha = TRUE,
      na.color = "transparent"
    )
  }
  
  
  ## grab the missing ecoregions
  missingEcos <- data$ersex$missingEcos |> unlist()
  ## filter the full set
  ecoReg <- data$naturalArea |>
    dplyr::filter(ECO_ID_U %in% missingEcos) |>
    dplyr::mutate(ECO_NAME = as.factor(ECO_NAME))
  
  # need to generate a raster object ecoregions, this should be done as part of the modeling process
  # then I can pass through a data frame of the ecoregion data, rather the spatial objects
  
  # Rasterize the ecoregions
  webMecEcoReg <- st_transform(ecoReg, crs = 3857)
  ecoRast <- terra::rasterize(x = webMecEcoReg, y = threshold, "ECO_ID_U")
  
  # apply the mask
  ecoRast <- mask(ecoRast, mask1)
  
  
  # set the palette
  # this defines the color used, and can be called in the legend creation
  ## set 3 has a cap of 12 unique colors
  ### duplicating the values then subsetting base on the required number.
  ecoPal <-  rep(RColorBrewer::brewer.pal(n = 12, name = "Spectral"), 6)
  if (length(ecoReg$ECO_ID_U) > 0) {
    ecoPal <- ecoPal[1:length(ecoReg$ECO_ID_U)]
    
    ecoReg$color <- ecoPal
    ecoReg <- ecoReg |> arrange(desc(ECO_NAME))
    ecoReg$ECO_ID_U <- factor(ecoReg$ECO_ID_U,
                              ordered = TRUE,
                              levels = c(ecoReg$ECO_ID_U))
    
    # define a palette option so that we can set a transparent NA value
    ecoColor <- colorFactor(
      palette = ecoReg$color,
      domain = ecoReg$ECO_ID_U,
      na.color =  "#FF000000"
    )
  } else{
    ecoColor <- "#FF000000"
  }
  
  
  
  #distribution palette
  distColor <- c("#ffffd4", "#fed98e", "#fe9929", "#d95f0e", "#993404")
  # protect area palette
  proColor <- c("#7570b3")
  # srsin palete
  srsPal1 <-  c("#cfcccc95",
                "#7fcdbb",
                "#41b6c4",
                "#1d91c0",
                "#225ea8",
                "#0c2c84")
  srsPal <- colorNumeric(
    palette = c(
      "#cfcccc95",
      "#7fcdbb",
      "#41b6c4",
      "#1d91c0",
      "#225ea8",
      "#0c2c84"
    ),
    domain = values(srs1),
    alpha = TRUE,
    na.color = "transparent"
  )
  
  # Mean distribution map
  meanDist <- data$projectedResults$mean
  meanPal <- colorNumeric(palette = distColor,
                          domain = values(meanDist),
                          na.color = "transparent")
  # Median distribution map
  medianDist <- data$projectedResults$median
  medianPal <-
    colorNumeric(
      palette = distColor,
      domain = values(medianDist),
      na.color = "transparent"
    )
  
  # standdard devation map
  stdevDist <- data$projectedResults$stdev
  stdevPal <- colorNumeric(
    palette = distColor,
    domain = values(stdevDist),
    na.color = "transparent"
  )
}

```


<br>

## Map of Predicted distribution

The map below shows the a 50km buffer area sounding all available points. This area was masked to the ecoregions in which a given occurrence data has been observed.


```{r echo=FALSE, message=FALSE, warning=FALSE}
# modeled distribution Map

# Add occurrances
m1 <- leaflet(width = "100%", options = leafletOptions(minZoom = 3)) |>
  addProviderTiles(providers$Esri.WorldGrayCanvas) |>
  addCircleMarkers(
    data = occData[occData$type == "H", ],
    color = ~ color,
    opacity = 1,
    radius = 1,
    group = "Occurrences",
    stroke = 1,
    popup = ~ popup,
    clusterOptions = markerClusterOptions()
  ) |>
  addCircleMarkers(
    data = occData[occData$type == "G", ],
    color = ~ color,
    opacity = 1,
    radius = 1,
    group = "Occurrences",
    stroke = 1,
    popup = ~ popup,
    clusterOptions = markerClusterOptions()
  ) |>
  leaflet::addLegend(
    position = "bottomright",
    color = c("#1184D4", "#6300F0"),
    labels = c("H (for herbaria) observations", "G (for genebank) samples"),
    title = "Observation Type",
    group = "Occurrences"
  )



if (data$NoModel == FALSE) {
  # Trim rasters to remove empty edge space and reduce HTML payload size
  threshold_trim <- raster::trim(raster(threshold))
  
  ## add rasters
  m2 <- m1 |>
    # threshold
    addRasterImage(
      x = threshold_trim,
      colors = thresColor,
      group = "Buffered Area",
      project = FALSE
    ) |>
    leaflet::addLegend(
      position = "bottomright",
      color = c("#cfcccc", "#45B320"),
      labels = c("Native Area", "Buffered Area"),
      title = "Buffered Habitat",
      group = "Buffered Area"
    )
  
  ## Add control layers
  m3 <- m2 |>
    addLayersControl(
      position = "topleft",
      options = list(collapsed = FALSE, autoZIndex = FALSE),
      overlayGroups = c(
        "Buffered Area",
        "Occurrences"
      )
    )
  m3
} else{
  m1
}


```

<br>




<br>
<br>

#  *Ex situ* conservation gap analysis 

The table below shows the *ex situ* conservation gap analysis summary. The Sampling Representativeness Score *ex situ* (SRS ex situ) calculates the ratio of germplasm accessions (G) available in *ex situ* repositories to reference/voucher (H) records for each taxon, making use of all compiled records, regardless of whether they include coordinates. The Geographical Representativeness Score *ex situ* (GRS ex situ) uses buffers of 50 km radius created around each G collection coordinate point to estimate geographic areas already well collected within the 50 km buffer areas of each taxon, and then calculates the proportion of the 50 km buffer area covered by these buffers. The Ecological Representativeness Score *ex situ* (ERS ex situ) calculates the proportion of terrestrial ecoregions represented within the G buffered areas out of the total number of ecoregions occupied by the 50 km buffer area. A Final Conservation Score for *ex situ* (FCS ex situ) was derived by calculating the average of the three *ex situ* conservation metrics. All of the conservation metrics are on a scale from 0-100, with 0 = poor state of conservation and 100 = comprehensive conservation. The FCS ex situ is used to categorize species, with urgent priority (UP) for further conservation action assigned when FCS ex situ < 25, high priority (HP) where 25 ≤ FCS ex situ < 50, medium priority (MP) where 50 ≤ FCS ex situ < 75, and low priority (LP) for taxa whose FCS ex situ ≥75.


```{r echo=FALSE, message=FALSE, warning=FALSE}
exCounts <- data$countsData |> 
  dplyr::mutate(across(c('species'), str_replace_all, pattern = "_", replacement = " "))|>
  dplyr::select("Taxon" = species, 
                "Total with coordinates" = totalUseful,
                "Total G records" = totalGRecords,
                "Total G with coordinates" = totalGUseful,
                "Total H records" = totalHRecords,
                "Total H with coordinates" = totalHUseful)
exScores <- data$fcsex |> 
  dplyr::mutate("SRS ex situ" = sigfig(SRS),
                "GRS ex situ" = sigfig(GRS),
                "ERS ex situ" = sigfig(ERS),
                "FCS ex situ" = sigfig(FCS))|>
  dplyr::select("SRS ex situ","GRS ex situ","ERS ex situ","FCS ex situ")|>
  dplyr::mutate(
    "FCS ex situ priority category" = case_when(
      `FCS ex situ` >= 75 ~ "LP",
      `FCS ex situ` >=50 & `FCS ex situ` < 75 ~ "MP",
      `FCS ex situ` >=25 & `FCS ex situ` < 50 ~ "HP",
      `FCS ex situ` < 25 ~ "UP"
    )
  )
# combine the data
exTables <- bind_cols(exCounts,exScores)
#visualize 
DT::datatable(exTables,
              rownames = FALSE,
              class = "compact",
              options = list(dom = 't'))

```
<br>

## Map of *ex situ* conservation gap analysis measures

The GRS ex situ map feature shows the 50 km buffer area, with previous *ex situ* germplasm collection points surrounded by a 50 km buffer overlaid. Only germplasm (G) points are displayed on the map.

The ERS ex situ map feature shows ecoregions within the 50 km buffer area from which no *ex situ* germplasm collections have been made. The ecoregion extent has been clipped to be within the 50 km buffer area of the taxon. There are many cases where the ecoregion area within the 50 km buffer area extent is very small (e.g., less than 5 cells). Utilize the “All ecoregions” layer and the ERS ex situ legend to help locate these areas.

Displayed maps, projected at 2.5 arc minutes resolution here, may not precisely reflect conservation gap analysis scores, which were based on maps at 30 arc seconds resolution.


```{r echo=FALSE, message=FALSE, warning=FALSE}

# start with a map of the distribution and ecoregions 

if(data$NoModel == FALSE){
  ex1 <- leaflet(width = "100%")|>
        addProviderTiles(providers$Esri.WorldGrayCanvas)|>
        # Outline of all ecoregions
        addPolygons(
          data = allEcos,
          group = "All ecoregions",
          weight = 2,
          stroke = TRUE,
          fill = TRUE,
          fillOpacity = 0.2,
          popup = allEcos$ECO_NAME,
          color = "#60616050"
        ) |>
      leaflet::addRasterImage(x = threshold_trim,
                   colors = thresColor, 
                   group = "Buffered Area",
                   project = FALSE)
} else {
  ex1 <- leaflet(width = "100%") |>
      addProviderTiles(providers$Esri.WorldGrayCanvas)
}


# Add the GRSex maps 
if(data$NoModel == FALSE){
  # with No G points avoid adding the point features 
  if(exTables$`Total G with coordinates` == 0){
        grs_html <- HTML(
      "<div style='
        background: white; 
        padding: 6px 8px; 
        border-radius: 5px; 
        box-shadow: 0 1px 5px rgba(0,0,0,0.4); 
        font-family: sans-serif;
      '>
        <strong>GRS Score: 0</strong> <br> No Map Elements
      </div>"
    )
    
    exG <- ex1|>
    leaflet::addLegend(position = "bottomright",
            color = c( "#cfcccc","#45B320"),
            labels = c("Native Area", "Buffered Area"),
            title = "Buffered Habitat",
            group = "Buffered Area")|>
              addControl(
    html = grs_html,
    position = "bottomleft"  # You can change this position
  )     
  }else{ # G points with valid lat lon are present
    # generate the GRSex map
    grsexMap_trim <- raster::trim(raster(grsexMap))
    exG <- ex1 |>
      #GRSex raster
      addRasterImage(
        x = grsexMap_trim,
        colors = grsexColor,
        group = "GRS ex situ",
        project = FALSE)   |>
      #G points
      addCircleMarkers(
        data = occData[occData$type == "G",],
        color = ~ color,
        opacity = 1,
        radius = 1,
        group = "GRS ex situ",
        stroke = 1,
        popup = ~ popup
      ) |>
      #add legend
      leaflet::addLegend(
        position = "bottomright",
        color = exPal,
        title = "Ex Situ geographical representativeness",
        labels = c("Native country-ecoregion area", "Buffered Area", "Areas < 50 km from an ex situ collection"),
        group = "GRS ex situ"
      )
  } 
}

# Add the ERSex maps 
# Add the ERSex maps 
if(data$NoModel == FALSE){
  # generate the ERSex map
  # SAFTEY CHECK: Catch both perfect scores AND 0-row aggregated tables
  if(data$fcsex$ERS == 100 || nrow(ecoReg) == 0){
    ers_html <- HTML(
      "<div style='
        background: white; 
        padding: 6px 8px; 
        border-radius: 5px; 
        box-shadow: 0 1px 5px rgba(0,0,0,0.4); 
        font-family: sans-serif;
      '>
        <strong>ERS Score: 100 </strong> <br> No Map Elements
      </div>"
    )

    exE <- exG |>
      addControl(
    html = ers_html,
    position = "bottomleft")
    } else {
      ecoRast_trim <- raster::trim(raster(ecoRast))
      exE <- exG |>
      # Ecoregion Raster Layers
      addRasterImage(ecoRast_trim,
                   group = "ERS ex situ",
                   colors = ecoColor,
                   project = FALSE) |>
      leaflet::addLegend(colors = ecoReg$color,
                group = "ERS ex situ",
                labels = ecoReg$ECO_NAME) |>
      # Outline of all ecoregions
      addPolygons(
        data = ecoReg,
        group = "Missing ecoregions",
        color = "#f0a01f",  # This sets the outline color
        weight = 0.5,         # Outline width
        opacity = 0.5,        # Outline opacity
        fillOpacity = 0,    # Set to 0 for no fill
        popup = ~ECO_NAME)
    }
  }
  
  # add controls for the layout
  ex3 <- exE |>
    addLayersControl(
      position = "topleft",
      options = list(collapsed = FALSE,
                     autoZIndex = FALSE),
      overlayGroups = c("Buffered Area",
        "GRS ex situ",
        "ERS ex situ",
        "Missing ecoregions",
        "All ecoregions"
      )
    ) |>
    hideGroup(c(
      "ERS ex situ"
    ))
  
  ## print the map
  ex3

```

<br>

### *Ex situ* conservation ecoregions table


The table below shows all the ecoregions overlapping with the 50 km buffer area of the taxon. The column ERS ex situ gap ecoregion denotes (true) if an ecoregion overlaps with the 50 km buffer area but is farther than 50 km from any *ex situ* conservation occurrence, and is thus considered a gap in *ex situ* conservation.

```{r echo=FALSE, message=FALSE, warning=FALSE}


# All ecoregions considered
# identify if they
## have buffered habitat
## are within a 50km buffer of a germplasm collection
if(data$NoModel == FALSE){
  eco2 <- data$naturalArea |>
  st_drop_geometry() |>
  dplyr::mutate(
    'ERSex Gap Ecoregion' = case_when(ECO_ID_U %in% missingEcos ~ TRUE,
                                      TRUE ~ FALSE)
  ) |>
  # dplyr::left_join(misArea, by = c("ECO_ID_U"="value"))|>
  dplyr::select(
    `ERS ex situ gap ecoregion` = 'ERSex Gap Ecoregion',
    # "Square KM missing" = area,
    `Unique ID` = "ECO_ID_U",
    `Ecoregion ame` = "ECO_NAME"  ) |>
  arrange(desc(`ERS ex situ gap ecoregion`))
#visualize
DT::datatable(eco2,
              rownames = FALSE,
              class = "compact")

}else{
  eco2 <- data$naturalArea |>
  st_drop_geometry() |>
  dplyr::mutate(
    'ERSex Gap Ecoregion' =  TRUE) |> 
  # dplyr::left_join(misArea, by = c("ECO_ID_U"="value"))|>
  dplyr::select(
    `ERS ex situ gap ecoregion` = 'ERSex Gap Ecoregion',
    # "Square KM missing" = area,
    `Unique ID` = "ECO_ID_U",
    `Ecoregion name` = "ECO_NAME"  ) |>
  arrange(desc(`ERS ex situ gap ecoregion`))
#visualize
  DT::datatable(eco2,
              rownames = FALSE,
              class = "compact")
}

``` 

<br>
<br>

# *In situ* conservation gap analysis

The table below shows the *in situ* conservation gap analysis summary. The Sampling Representativeness Score *in situ* (SRS in situ) calculates the proportion of all occurrences of a taxon within its buffered area that fall within a protected area. The Geographical Representativeness Score *in situ* (GRS in situ) compares the area (in km^2^) of the 50 km buffer area located within protected areas versus the total buffered area. The Ecological Representativeness Score *in situ* (ERS in situ) calculates the proportion of ecoregions encompassed within the range of the taxon located inside protected areas to the ecoregions encompassed within the total area of the 50 km buffer area. A Final Conservation Score for *in situ* (FCS in situ) was derived by calculating the average of the three *in situ* conservation metrics. All of the conservation metrics are on a scale from 0-100, with 0 = poor state of conservation and 100 = comprehensive conservation. The FCS in situ is used to categorize taxa, with urgent priority (UP) for further conservation action assigned when FCS in situ < 25, high priority (HP) where 25 ≤ FCS in situ < 50, medium priority (MP) where 50 ≤ FCS in situ < 75, and low priority (LP) for taxa whose FCS in situ ≥75.

```{r echo=FALSE, message=FALSE, warning=FALSE}

inCounts <- data$countsData |> 
  dplyr::mutate(across(c('species'), str_replace_all, pattern = "_", replacement = " "))|>
  dplyr::mutate("Total occurrences in protected areas" = round(data$countsData$totalUseful * (data$fcsin$SRS/100))) |> 
  dplyr::select("Taxon" = species, 
                "Total occurrences" = totalUseful,
                  "Total occurrences in protected areas")
if(data$NoModel == FALSE){
  inScores <- data$fcsin |> 
  dplyr::mutate("SRS in situ" = sigfig(SRS),
                "GRS in situ" = sigfig(GRS),
                "ERS in situ" = sigfig(ERS),
                "FCS in situ" = sigfig(FCS),
                "Total occurrences in buffered area" = totalObsinThres,
                "Total occurrences in protected areas" = totalObsinPro)|>
  dplyr::select("SRS in situ","GRS in situ","ERS in situ","FCS in situ")|>
  dplyr::mutate(
    "FCS in situ priority category" = case_when(
      `FCS in situ` >= 75 ~ "LP",
      `FCS in situ` >=50 & `FCS in situ` < 75 ~ "MP",
      `FCS in situ` >=25 & `FCS in situ` < 50 ~ "HP",
      `FCS in situ` < 25 ~ "UP"
    )
  )
# combine the data
inTables <- bind_cols(inCounts,inScores)
#visualize 
DT::datatable(inTables,
              rownames = FALSE,
              class = "compact",
              options = list(dom = 't'))
}else{
    inScores <- data$fcsin |> 
  dplyr::mutate("SRS in situ" = sigfig(SRS),
                "GRS in situ" = sigfig(GRS),
                "ERS in situ" = sigfig(ERS),
                "FCS in situ" = sigfig(FCS),
                "Total occurrences in buffered area" = NA,
                "Total occurrences in protected areas" = inCounts$`Total occurrences in protected areas`)|>
  dplyr::select("SRS in situ","GRS in situ","ERS in situ","FCS in situ")|>
  dplyr::mutate(
    "FCS in situ priority category" = case_when(
      `FCS in situ` >= 75 ~ "LP",
      `FCS in situ` >=50 & `FCS in situ` < 75 ~ "MP",
      `FCS in situ` >=25 & `FCS in situ` < 50 ~ "HP",
      `FCS in situ` < 25 ~ "UP"
    )
  )
# combine the data
inTables <- bind_cols(inCounts,inScores)
#visualize 
DT::datatable(inTables,
              rownames = FALSE,
              class = "compact",
              options = list(dom = 't'))
}



```

<br>
<br>

## Map of *in situ* conservation gap analysis measures

<br>

The SRS in situ map feature shows occurrences within the 50km buffer that are outside of protected areas. Values are summarized by count per cell. The GRS in situ map feature shows the 50km buffer, with buffered area occurring within existing protected areas in purple and that outside of protected areas in green. The ERS in situ map features shows ecoregions within the 50 km buffer area that have no protected areas. There are many cases where the ecoregion area within the 50 km buffer area extent is very small (e.g., less than 5 cells). Utilize the “All ecoregions” layer and the ERS in situ legend to help locate these areas.

Displayed maps, projected at 2.5 arc minutes resolution here, may not precisely reflect conservation gap analysis scores, which were based on maps at 30 arc seconds resolution.



```{r echo=FALSE, message=FALSE, warning=FALSE}


# generate the base map 
if(data$NoModel == FALSE){
  in1 <- leaflet(width = "100%")|>
        addProviderTiles(providers$Esri.WorldGrayCanvas)|>
        # Outline of all ecoregions
        addPolygons(
          data = allEcos,
          group = "All ecoregions",
          weight = 2,
          stroke = TRUE,
          fill = TRUE,
          fillOpacity = 0.2,
          popup = allEcos$ECO_NAME,
          color = "#60616050"
        ) |>
      leaflet::addRasterImage(x = threshold_trim,
                   colors = thresColor, 
                   group = "Buffered Area",
                   project = FALSE)
} else {
  in1 <- leaflet(width = "100%") |>
      addProviderTiles(providers$Esri.WorldGrayCanvas)
}

# srsin map 
if(data$NoModel == FALSE){
  uniqueSRSVals <- sort(unique(values(srs1)))
  # condition for no values 
  if(length(uniqueSRSVals) == 1){
    srs_html <- HTML(
      "<div style='
        background: white; 
        padding: 6px 8px; 
        border-radius: 5px; 
        box-shadow: 0 1px 5px rgba(0,0,0,0.4); 
        font-family: sans-serif;
      '>
        <strong>SRS Score: 0 </strong> <br> No Map Elements
      </div>"
    )

    inS <- in1|>
      addControl(
    html = srs_html,
    position = "bottomleft")
  }else{

srs1_trim <- raster::trim(raster(srs1))
protectedArea_trim <- raster::trim(raster(protectedArea))

inS <- in1 |> 
  # SRSin 
  addRasterImage(
    x = srs1_trim,
    colors = srsPal,
    group = "SRS in situ",
    project = FALSE)|>
  # protected Areas 
  addRasterImage(
    x = protectedArea_trim,
    colors = "#7570b3",
    group = "SRS in situ",
    project = FALSE )|>
  leaflet::addLegend(colors = srsPal1,
            group = "SRS in situ",
            title = "SRS in situ",
            labels = c(min(uniqueSRSVals),"", "","","",max(uniqueSRSVals)))|>
  leaflet::addLegend(colors = "#7570b3",
            group = "SRS in situ",
            # title = "Protected Areas",
            labels = c("Protected Areas"))
  }
}

# GRS map 
if(data$NoModel == FALSE){
uniqueSRSVals <- sort(unique(values(srs1)))
  # condition for no values 
  if(data$fcsin$GRS == 0){
    grs_html <- HTML(
      "<div style='
        background: white; 
        padding: 6px 8px; 
        border-radius: 5px; 
        box-shadow: 0 1px 5px rgba(0,0,0,0.4); 
        font-family: sans-serif;
      '>
        <strong>GRS Score: 0 </strong> <br> No Map Elements
      </div>"
    )

    inG <- inS|>
      addControl(
    html = grs_html,
    position = "bottomleft")
  }else{

  ##GRSin map 
  grs1_trim <- raster::trim(raster(grs1))
  inG <- inS |>
    addRasterImage(
      x = grs1_trim,
      colors = grsinColor,
      group = "GRS in situ",
      project = FALSE
    )|>
    leaflet::addLegend(colors = exPal,
              group = "GRS in situ",
              labels = c("Background area", "Buffered Area", "Buffered Area within a protected area"))
  }
}

# ERS map 
if(data$NoModel == FALSE){
  # condition for no values 
  if(data$fcsin$ERS == 100){
    ers_html <- HTML(
      "<div style='
        background: white; 
        padding: 6px 8px; 
        border-radius: 5px; 
        box-shadow: 0 1px 5px rgba(0,0,0,0.4); 
        font-family: sans-serif;
      '>
        <strong>ERS Score: 100 </strong> <br> No Map Elements
      </div>"
    )

    inE <- inG|>
      addControl(
    html = ers_html,
    position = "bottomleft")
  }else{
  missEcoIn <- allEcos[allEcos$ECO_ID_U %in% ersin$missingEcos, ]
  ##ERSin map 
  ersinMap_trim <- raster::trim(raster(ersinMap))
  inE <- inG |>
  addRasterImage(x = ersinMap_trim,
    colors = ecoInColor,
    group = "ERS in situ",
    project = FALSE
  )|>
    leaflet::addLegend(colors = ers2$color,
              group = "ERS in situ",
              labels = ers2$ECO_NAME)|>
                      # Outline of all ecoregions
      addPolygons(
        data = missEcoIn,
        group = "Missing ecoregions",
        color = "#f0a01f",  # This sets the outline color
        weight = 0.5,         # Outline width
        opacity = 0.5,        # Outline opacity
        fillOpacity = 0,    # Set to 0 for no fill
        popup = ~ECO_NAME)
  }
}

# add control groups 
inMap <- inE |> 
  addLayersControl(
    position = "topleft",
    options = list(collapsed = FALSE,
                   autoZIndex = FALSE),
    overlayGroups = c(
      "Buffered Area",
      "SRS in situ",
      "GRS in situ",
      "ERS in situ",
      "Missing ecoregions",
      "All ecoregions"))|>
  hideGroup(group = c( "SRS in situ", "ERS in situ","All ecoregions"))

inMap

```
<br>

### *In situ* conservation ecoregions table

The table below shows all the ecoregions overlapping with the 50 km buffer area of the taxon. The column ERS in situ gap ecoregion denotes (true) if an ecoregion overlaps with the 50 km buffer area but none of that area is within a protected area, and is thus considered a gap in *in situ* conservation.



```{r echo=FALSE, message=FALSE, warning=FALSE}
# All ecoregions considered
# identify if they
## have buffered habitat
## are within a 50km buffer of a germplasm collection
if(data$NoModel == FALSE){
  eco3 <- data$naturalArea |>
  st_drop_geometry() |>
  dplyr::mutate(
    'ERS in situ gap ecoregion' = case_when(ECO_ID_U %in% ers2$ECO_ID_U ~ TRUE,
                                      TRUE ~ FALSE)
  ) |>
  # dplyr::left_join(misAreaIn, by = c("ECO_ID_U"= "values"))|>
  dplyr::select(
    'ERS in situ gap ecoregion',
    # 'Square KM missing' = values,
    `Unique ID` = "ECO_ID_U",
    `Ecoregion name` = "ECO_NAME",
  ) |>
  arrange(desc(`ERS in situ gap ecoregion`))
#visualize
DT::datatable(eco3,
              rownames = FALSE,
              class = "compact")
}else{
  "No model was generated for this species"
}


``` 



<br>

<br>

<br>

# Combined conservation gap analysis

The table below displays all conservation gap analysis scores, providing a comparison of the *ex situ* and *in situ* conservation scores.
```{r echo=FALSE, message=FALSE, warning=FALSE,fig.align='right',out.width="100%"}
exTable <- data$fcsex|>
  dplyr::mutate(across(c('ID'), str_replace_all, pattern = "_", replacement = " "))
names(exTable) <- c("Taxon", "Ex situ Sampling Representativeness Score", "Ex situ Geographic Representativeness Score", "Ex situ Ecological Representativeness Score", "Ex situ Final Conservation Score", "Ex situ Conservation Priority")
inTable <- data$fcsin |>
  dplyr::select(- ID)
names(inTable) <- c("In situ Sampling Representativeness Score",
                    "In situ Geographic Representativeness Score",
                    "In situ Ecological Representativeness Score",
                    "In situ Final Conservation Score",
                    "In situ Conservation Priority")
  
finalTable <- bind_cols(exTable, inTable)|>
  # round values in specific column 
  dplyr::mutate(across(c(2:5,7:10), round, digits = 2))
 


# generate the combine conservation score 
meanVal <- mean(c(finalTable$`Ex situ Final Conservation Score`, finalTable$`In situ Final Conservation Score`))

finalTable <- finalTable |>
  dplyr::mutate(
    `Final Conservation Score Mean` = round(meanVal, digits = 2),
    `Combined Conservation Priority` = case_when(
      meanVal <= 25 ~ "UP",
      meanVal > 25 & meanVal <= 50 ~ "HP",
      meanVal > 50 & meanVal <= 75 ~ "MP",
      meanVal <= 75 ~ "UP",
    )
  )
#visualize
DT::datatable(finalTable,
              rownames = FALSE,
              class = "compact",
              options = list(dom = 't'))

```
<br>

## Conservation scores 

The figure below displays all conservation gap analysis scores, providing a comparison of the *ex situ* and *in situ* conservation scores.

```{r echo=FALSE, message=FALSE, warning=FALSE,fig.align='right',out.width="100%"}
# Data preparation
d2 <- rbind(data$fcsex, data$fcsin) |>
  mutate(Category = c("ex situ", "in situ")) |>
  dplyr::select(
    ID,
    "SRS" = SRS,
    "GRS" = GRS,
    "ERS" = ERS,
    "FCS" = FCS,
    "Score Class" = FCS_Score,
    Category
  )

d3 <- d2 |>
  pivot_longer(cols = c("SRS", "GRS", "ERS", "FCS"), names_to = "Metric")

# Define the combined score for the species 
combined_mean <- as.numeric(mean(c(d3$value[4], d3$value[8])))

d3[9, ] <- list(
  d3$ID[1], 
  NA, 
  "Combined conservation score", 
  "Combined Final<br>Conservation Score", 
  combined_mean
)

# Define the combined score category 
if (d3$value[9] <= 25) {
  d3$`Score Class`[9] <- "UP"
} else if (d3$value[9] <= 50) {
  d3$`Score Class`[9] <- "HP"
} else if (d3$value[9] <= 75) {
  d3$`Score Class`[9] <- "MP"
} else {
  d3$`Score Class`[9] <- "LP"
}

# Update factor levels to match the new string with the line break
d3$Metric <- factor(
  d3$Metric, 
  levels = c("SRS", "GRS", "ERS", "FCS", "Combined Final<br>Conservation Score")
)

# Plotting
plot_ly(d3,
        x = ~Metric,
        y = ~value,
        color = ~Category) |>
  layout(
    xaxis = list(
      title = "", 
      tickangle = 0, 
      
      # --- Add these two lines to force the order ---
      categoryorder = "array",
      categoryarray = c("SRS", "GRS", "ERS", "FCS", "Combined Final<br>Conservation Score"),
      # ----------------------------------------------
      
      tickmode = "array",
      tickvals = c("SRS", "GRS", "ERS", "FCS", "Combined Final<br>Conservation Score"),
      ticktext = c("              SRS", "              GRS", "              ERS", "              FCS", "Combined Final<br>Conservation Score") 
    ), 
    yaxis = list(
      title = "", 
      range = c(0, 100) # Locks the y-axis to always show 0 to 100
    ),
    margin = list(b = 60), 
    barmode = 'group',
    bargroupgap = 0, # Pushes the bars together
    hovermode = 'compare'
  )
```

*Note: Score values of zero will not appear on the chart.*


## Definitions of conservation gap analysis scores

### Sampling Representativeness Score (SRS)

***Ex situ***: The Sampling Representativeness Score *ex situ* (SRS ex situ) calculates the ratio of germplasm accessions (G) available in *ex situ* repositories to reference/voucher/herbarium (H) records for each taxon, making use of all compiled records irrespective of whether they include coordinates.

***In situ***: The Sampling Representativeness Score *in situ* (SRS in situ) calculates the proportion of all occurrences of a taxon within its native range that fall within a protected area.

### Geographic Representativeness Score (GRS)

***Ex situ***: The Geographic Representativeness Score *ex situ* (GRS ex situ) uses 50-km-radius buffers created around each G collection coordinate point to estimate geographic areas already well collected within the 50 km buffer areas of each taxon and then calculates the proportion of the 50 km buffer area covered by these buffers.

***In situ***: The Geographic Representativeness Score *in situ* (GRS in situ) compares the area (in km^2^) of the 50 km buffer area located within protected areas versus the total buffered area.

### Ecological Representativeness Score (ERS)

***Ex situ***: The Ecological Representativeness Score *ex situ* (ERS ex situ) calculates the proportion of terrestrial ecoregions represented within the G buffered areas out of the total number of ecoregions occupied by the 50 km buffer area. 

***In situ***: The Ecological Representativeness Score *in situ* (ERS in situ) calculates the proportion of ecoregions encompassed within the range of the taxon located inside protected areas to the ecoregions encompassed within the total area of the 50 km buffer area.

### Final Conservation Score (FCS)

***Ex situ***: The Final Conservation Score *ex situ* (FCS ex situ) was derived by calculating the average of the three *ex situ* conservation metrics.

***In situ***: The Final Conservation Score *in situ* (FCS in situ) was derived by calculating the average of the three *in situ* conservation metrics.

**FCSc-mean**: The Combined Final Conservation Score (FCSc-mean) was calculated for each taxon by averaging its final FCS ex situ and FCS in situ scores. 

Taxa were further categorized with regard to the two conservation strategies as well as in combination, with Urgent Priority (UP) for further conservation action assigned when FCS < 25, High Priority (HP) assigned when 25 ≤ FCS < 50, Medium Priority (MP) when 50 ≤ FCS < 75, and Low Priority (LP) when FCS ≥75.

## Definitions of occurrence data categories

**Occurrences**: The total number of observations and *ex situ* collections records for the taxon that were evaluated for use within the study.

**Occurrences with lat/long**: The total number of observations that had a valid latitude and longitude pair and were used for spatial analysis. Non-valid latitude longitude pairs included but were not limited to points not on land masses, and records with only one of the two values present.

**Germplasm Records (G)**: Occurrences in which a living sample (typically a live plant or seed) is present in an *ex situ* conservation system (i.e., botanical garden, seed bank, genebank, etc.).


**Reference/voucher Records (H)**: Occurrences that have a supporting herbarium or other reference record.

**Unique Data Sources**: The number of unique database sources from which occurrences of the taxon were gathered.


