---
title: Taxon level predicted distribution modeling and conservation gap analysis summary
author: Daniel P. Carver, Colin K. Khoury, Sarah L. Gora
output:
  html_document:
    css: style.css
date: "Report generated on ```r Sys.Date()```"
params:
  reportData: NA
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
pacman::p_load(
  "leaflet",
  "data.table",
  "tidyr",
  "dplyr",
  "plotly",
  "DT",
  "RColorBrewer",
  "htmltools"
)

### function for control number of character after the decimal place
sigfig <- function(vec, n = 3) {
  ### function to round values to N significant digits
  # input:   vec       vector of numeric
  #          n         integer is the required sigfig
  # output:  outvec    vector of numeric rounded to N sigfig
  
  formatC(
    signif(vec, digits = n),
    digits = n,
    format = "fg",
    flag = "#"
  )
  
}  
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
testing <- FALSE
# testing <- TRUE
if(testing == TRUE){
  data <- reportData # testing only
  species <- data$occuranceData$taxon[1] # testing
}else{
  data <- params$reportData
  species <- data$occuranceData$taxon[1]
}

printSpecies <- stringr::str_replace_all(string = species, pattern = "_", replacement = " ")

# --- Reduce polygon complexity and coordinate precision to reduce HTML size ---
round_sf_coords <- function(sf_obj, digits = 3) {
  if (is.null(sf_obj) || nrow(sf_obj) == 0) return(sf_obj)
  geom <- sf::st_geometry(sf_obj)
  geom <- sf::st_sfc(lapply(geom, function(x) {
    if (inherits(x, "POLYGON")) {
      sf::st_polygon(lapply(x, function(y) round(y, digits)))
    } else if (inherits(x, "MULTIPOLYGON")) {
      sf::st_multipolygon(lapply(x, function(y) lapply(y, function(z) round(z, digits))))
    } else if (inherits(x, "POINT")) {
      sf::st_point(round(as.numeric(x), digits))
    } else {
      x
    }
  }), crs = sf::st_crs(sf_obj))
  sf::st_geometry(sf_obj) <- geom
  return(sf_obj)
}

if (!is.null(data$naturalArea)) {
  # Simplify polygons predictably by turning off S2 temporarily 
  # (so dTolerance is in degrees: 0.05 deg is ~5.5km)
  s2_state <- sf::sf_use_s2()
  sf::sf_use_s2(FALSE)
  
  is_longlat <- sf::st_is_longlat(data$naturalArea)
  tol <- if (!is.na(is_longlat) && is_longlat) 0.05 else 5000
  data$naturalArea <- sf::st_simplify(data$naturalArea, preserveTopology = FALSE, dTolerance = tol)
  
  sf::sf_use_s2(s2_state)
  
  # Round coordinates (2 digits is ~1.1km precision)
  digits_to_keep <- if (!is.na(is_longlat) && is_longlat) 2 else 0
  data$naturalArea <- round_sf_coords(data$naturalArea, digits = digits_to_keep)
}

if (!is.null(data$occuranceData)) {
  if (inherits(data$occuranceData, "sf")) {
    is_longlat <- sf::st_is_longlat(data$occuranceData)
    digits_to_keep <- if (!is.na(is_longlat) && is_longlat) 3 else 0
    data$occuranceData <- round_sf_coords(data$occuranceData, digits = digits_to_keep)
  }
  if ("longitude" %in% names(data$occuranceData)) {
    data$occuranceData$longitude <- round(data$occuranceData$longitude, 3)
    data$occuranceData$latitude <- round(data$occuranceData$latitude, 3)
  }
}
# ----------------------------------------------------------------------------


# remove the g point from MBG for bloodworthiana 
if(species == "Vitis bloodworthiana"){
  data$occuranceData <- data$occuranceData[data$occuranceData$longitude != -100.00000, ]
}


```

# `r paste0('*', printSpecies, '*')`
Occurrence records
The table below shows a summary of the occurrence records for the taxon. A record “with Lat/Long” is one that has a complete set of latitude and longitude values associated with it, which could be used for distribution modeling and the conservation gap analysis. In preparation for the conservation gap analyses, we classified each record based on whether it was an existing ex situ germplasm sample from a genebank, botanic garden, or other repository (labeled G, as most records were from genebanks), or a reference observation/voucher specimen (labeled H, as most of these records were from herbaria).

```{r echo=FALSE, message=FALSE, warning=FALSE}

# alter Counts data for better visualizations
c1 <- data$countsData |>
dplyr::select(
Taxon = species,
'Occurrences' = totalRecords,
'Occurrences with Lat/Long' = totalUseful,
'Germplasm Records(G)' = totalGRecords,
'Germplasm Records(G) with Lat/Long' = totalGUseful,
'Reference Records(H)' = totalHRecords,
'Reference Records(H) with Lat/Long' = totalHUseful,
'Unique Data Sources' = numberOfUniqueSources
) |>
dplyr::mutate(Taxon = stringr::str_replace_all(Taxon, pattern = "_", replacement = " "))

DT::datatable(c1,
rownames = FALSE,
class = "compact",
options = list(dom = 't'))

```

```{r, echo=FALSE, eval=TRUE}

# resample the raster datasets to 5km for visualization
# threshold maps
thresAgg <- matrix(c(
-Inf, 0.5, 0,  # All values from -Infinity up to (but not including) 0.5 become 0
0.5,  Inf, 1   # All values from 0.5 (including 0.5) up to +Infinity become 1
), ncol = 3, byrow = TRUE)

resampleFactor <- 5

resampleRast <- function(rast, resampleFactor = 5){
r1 <- rast |>
terra::aggregate(fact = resampleFactor, fun = "mean")  |>
terra::classify(thresAgg, include.lowest = TRUE)
return(r1)
}

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Generate all the map data
# point data
# Define a much shorter popup text to prevent Base64/HTML string bloat
occData <- data$occuranceData |>
dplyr::mutate(
popup = paste0(
"", taxon, "",
type, " (", databaseSource, ")"
)
) |>
mutate(color = case_when(type == "H" ~ "#1184D4",
type == "G" ~ "#6300F0"))

webMecOccData <- vect(st_transform(occData, crs = "epsg:3857")) # need to change project

if(dataNoModel == FALSE){
  # predicted presence map
  threshold <- databinaryMap
  
  thresColor <- colorNumeric(
  palette = c("#cfcccc70", "#45B320"),
  domain = values(threshold),
  alpha = TRUE,
  na.color = "transparent"
)

# generate a threshold mask
mask1 <- ifel(test = threshold == 1, yes = 1, no = NA)

# protected area
protectedArea <- data$protectedArea

protectColor <- colorNumeric(
palette = "#B7B648",
domain = values(protectedArea),
alpha = TRUE,
na.color = "transparent"
)

# protected areas mask
proMask <- ifel(test = protectedArea == 1, yes = 1, no = NA)

# insitu map elements
# SRSIN
# total number of observations per cell of the threshold model masked against the protect areas layer
# filter points outside of protected areas
webMecOccData <- vect(st_transform(occData, crs = "epsg:3857")) # need to change project

# for the spatial operations with the layer. occData can't have the proj
# get all the points that are outside of a protected area
inThres <- terra::extract(x = mask1, y = webMecOccData)|>
dplyr::filter(Threshold==1)
inPro <- terra::extract(x = protectedArea, y = webMecOccData) |>
dplyr::filter(is.na(layer))

srs1 <- terra::rasterize(webMecOccData,
threshold,
fun = "sum",
background = 0) |>
mask(mask1)

totalObsinThres <- nrow(inThres)

# total number of observations in protected areas
totalObsinPro <- nrow(inThres) - nrow(inPro)

# GRSIN
# reclassy protected area
proReclass <- ifel(test = is.na(protectedArea),
yes = 0,
no = 1)

# add values
grs1 <- threshold + proReclass
exPal <- c("#cfcccc80", "#45b320", "#7570b3")

# for some reason passing the vector exPal to the palette arg does not maintain the alpha reference?
grsinColor <- colorNumeric(palette = c("#cfcccc80", "#45b320", "#7570b3"),
# purple is a little lighter so the point locations show up clearly
domain = values(grs1),
alpha = TRUE,
na.color = "transparent")

# ERSin
# used on the ERS visualization
allEcos <- data$naturalArea |>
dplyr::select("ECO_ID_U", "ECO_NAME")

# reproject for spatial operations
webMecAllEcos <- st_transform(allEcos, crs = "epsg:3857")

# showcase ecoregions within the threshold model that do no have a protected area
ers1 <- terra::zonal(proReclass,
vect(webMecAllEcos),
fun = "sum",
na.rm = TRUE)
ers1$ECO_ID_U <− allEcos$ECO_ID_U
ers1$ECONAME <− allEcos$ECO_NAME

# zonal Stats to determine overlap between the ecoregions and the threshold model
v1 <- terra::zonal(x = threshold,
z = vect(webMecAllEcos),
fun = "sum",
na.rm = TRUE)# , exact = TRUE)

# reassign the eco id record
v1$ECO_ID_U  <− webMecAllEcos$ECO_ID_U

# get a list of ecoregions within the threshold model
ecosInThres <- v1 |>
dplyr::filter(Threshold > 0) |>
dplyr::select(ECO_ID_U) |>
pull()

# ID ecoregions in threshold model with no protected areas
ers2 <- ers1 |>
dplyr::filter(ECO_ID_U %in% ecosInThres) |>
dplyr::filter(layer == 0)

if(nrow(ers2) > 0) {
## use this list to generate a raster object masked to the
ersinMap <- webMecAllEcos |>
dplyr::filter(ECO_ID_U %in% ers2ECO_ID_U) |>
terra::rasterize(y = threshold, "ECO_ID_U") |>
mask(mask = mask1)
# generate palatte object for map 
ecoPal2 <-  rep(RColorBrewer::brewer.pal(n = 12, name = "Spectral"), 6)
ecoPal2 <- ecoPal2[1:length(ers2ECO_ID_U)]
# need to order the names so they appear as expected
ers2$color <− ecoPal2
ers2 <− ers2
  ∣>arrange(desc(ECONAME))

ers2color <- ecoPal2

ers2$ECO_ID_U <- factor(ers2$ECO_ID_U , ordered = TRUE, 
                                levels = c(ers2$ECO_ID_U))


# define a palette option so that we can set a transparent NA value
ecoInColor <- colorFactor(palette =ers2$color,
                          domain = ers2$ECO_ID_U,
                          na.color =  "#FF000000")
} else{
ersinMap <- "No values"
}

exsitu map elements
#buffered G area
gBuf <- data$g_bufferCrop

if (class(gBuf) != "character" & datafcsexGRS != 0) {

resample
gBuf <- gBuf |>
resampleRast()

bufColor <-
colorNumeric(
palette = "#6300F0",
domain = values(gBuf),
na.color = "transparent"
)

GRSex map element
m1 <- cbind(NaN, 0)
gbuf2 <- terra::classify(x = gBuf, m1)
grsexMap <- threshold + gbuf2

need to defined this outside of the palette function so it can be called again in the legend
grsexColor <- colorNumeric(palette = c("#cfcccc80", "#45b320", "#7570b3"),
# purple is a little lighter so the point locations show up clearly
domain = values(grsexMap),
alpha = TRUE,
na.color = "transparent")
}

#ecoregions

grab the missing ecoregions
missingEcos <- dataersexmissingEcos |> unlist()

filter the full set
ecoReg <- data$naturalArea |>
dplyr::filter(ECO_ID_U %in% missingEcos) |>
dplyr::mutate(ECO_NAME = as.factor(ECO_NAME))

need to generate a raster object ecoregions, this should be done as part of the modeling process
then I can pass through a data frame of the ecoregion data, rather the spatial objects
Rasterize the ecoregions
webMecEcoReg <- st_transform(ecoReg,crs = 3857)
ecoRast <- terra::rasterize(x = webMecEcoReg, y = threshold, "ECO_ID_U")

apply the mask
ecoRast <- mask(ecoRast, mask1)

set the palette
this defines the color used, and can be called in the legend creation
set 3 has a cap of 12 unique colors
duplicating the values then subsetting base on the required number.
ecoPal <-  rep(RColorBrewer::brewer.pal(n = 12, name = "Spectral"), 6)
if(length(ecoRegECO_ID_U)>0){
ecoPal <- ecoPal[1:length(ecoRegECO_ID_U)]

ecoRegcolor<−ecoPalecoReg<−ecoReg∣>arrange(desc(ECO 
N
​
 AME))ecoRegECO_ID_U <- factor(ecoRegECO 
I
​
 D 
U
​
 ,ordered=TRUE,levels=c(ecoRegECO_ID_U))

define a palette option so that we can set a transparent NA value
ecoColor <- colorFactor(palette = ecoRegcolor,domain=ecoRegECO_ID_U,
na.color =  "#FF000000")
}else{
ecoColor <- "#FF000000"
}

#distribution palette
distColor <- c("#ffffd4", "#fed98e", "#fe9929", "#d95f0e", "#993404")

protect area palette
proColor <- c("#7570b3")

srsin palete
srsPal1 <-  c("#cfcccc95",
"#7fcdbb",
"#41b6c4",
"#1d91c0",
"#225ea8",
"#0c2c84")
srsPal <- colorNumeric(palette = c("#cfcccc95","#7fcdbb","#41b6c4","#1d91c0","#225ea8","#0c2c84"),
domain = values(srs1),
alpha = TRUE,
na.color = "transparent")

}

```

Map of Predicted distribution
The map below shows the a 50km buffer area sounding all available points. This area was masked to the ecoregions in which a given occurrence data has been observed.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# modeled distribution Map
# Add occurrances
m1 <- leaflet(width = "100%", options = leafletOptions(minZoom = 3)) |>
addProviderTiles(providersEsri.WorldGrayCanvas)∣>addCircleMarkers(data=occData[occDatatype=="H",],
color = ~color,
opacity = 1,
radius = 1,
group = "Occurrences",
stroke = 1,
popup = ~popup,
clusterOptions = markerClusterOptions()) |>
addCircleMarkers(data = occData[occData$type=="G",],
color = ~color,
opacity = 1,
radius = 1,
group = "Occurrences",
stroke = 1,
popup = ~popup,
clusterOptions = markerClusterOptions())|>
leaflet::addLegend(
position = "bottomright",
color = c("#1184D4","#6300F0"),
labels = c("H (for herbaria) observations", "G (for genebank) samples"),
title = "Observation Type",
group = "Occurrences")

if(data$NoModel == FALSE){

# add rasters
m2 <- m1 |>
# threshold
addRasterImage(x = raster(threshold),
colors = thresColor,
group = "Buffered Area",
project = FALSE)|>
leaflet::addLegend(position = "bottomright",
color = c( "#cfcccc","#45B320"),
labels = c("Native Area", "Buffered Area"),
title = "Modeled Habitat",
group = "Buffered Area")

# Add control layers
m3 <- m2 |>
addLayersControl(position = "topleft", options = list(collapsed = FALSE,
autoZIndex = FALSE),
overlayGroups = c("Buffered Area","Occurrences"))
m3
}else{
m1
}

```

Ex situ conservation gap analysis
The table below shows the ex situ conservation gap analysis summary. The Sampling Representativeness Score ex situ (SRS ex situ) calculates the ratio of germplasm accessions (G) available in ex situ repositories to reference/voucher (H) records for each taxon, making use of all compiled records, regardless of whether they include coordinates. The Geographical Representativeness Score ex situ (GRS ex situ) uses buffers of 50 km radius created around each G collection coordinate point to estimate geographic areas already well collected within the distribution models of each taxon, and then calculates the proportion of the distribution model covered by these buffers. The Ecological Representativeness Score ex situ (ERS ex situ) calculates the proportion of terrestrial ecoregions represented within the G buffered areas out of the total number of ecoregions occupied by the distribution model. A Final Conservation Score for ex situ (FCS ex situ) was derived by calculating the average of the three ex situ conservation metrics. All of the conservation metrics are on a scale from 0-100, with 0 = poor state of conservation and 100 = comprehensive conservation. The FCS ex situ is used to categorize species, with urgent priority (UP) for further conservation action assigned when FCS ex situ < 25, high priority (HP) where 25 ≤ FCS ex situ < 50, medium priority (MP) where 50 ≤ FCS ex situ < 75, and low priority (LP) for taxa whose FCS ex situ ≥75.

```{r echo=FALSE, message=FALSE, warning=FALSE}
exCounts <- datacountsData∣>dplyr::mutate(across(c( 
′
 species 
′
 ),str 
r
​
 eplace 
a
​
 ll,pattern=" 
"
​
 ,replacement=""))∣>dplyr::select("Taxon"=species,"Totalwithcoordinates"=totalUseful,"TotalGrecords"=totalGRecords,"TotalGwithcoordinates"=totalGUseful,"TotalHrecords"=totalHRecords,"TotalHwithcoordinates"=totalHUseful)exScores<−datafcsex |>
dplyr::mutate("SRS ex situ" = sigfig(SRS),
"GRS ex situ" = sigfig(GRS),
"ERS ex situ" = sigfig(ERS),
"FCS ex situ" = sigfig(FCS))|>
dplyr::select("SRS ex situ","GRS ex situ","ERS ex situ","FCS ex situ")|>
dplyr::mutate(
"FCS ex situ priority category" = case_when(
FCS ex situ >= 75 ~ "LP",
FCS ex situ >=50 & FCS ex situ < 75 ~ "MP",
FCS ex situ >=25 & FCS ex situ < 50 ~ "HP",
FCS ex situ < 25 ~ "UP"
)
)

combine the data
exTables <- bind_cols(exCounts,exScores)
#visualize
DT::datatable(exTables,
rownames = FALSE,
class = "compact",
options = list(dom = 't'))

```


Map of ex situ conservation gap analysis measures
The GRS ex situ map feature shows the potential distribution model, with previous ex situ germplasm collection points surrounded by a 50 km buffer overlaid. Only germplasm (G) points are displayed on the map.

The ERS ex situ map feature shows ecoregions within the potential distribution model from which no ex situ germplasm collections have been made. The ecoregion extent has been clipped to be within the potential distribution model of the taxon. There are many cases where the ecoregion area within the potential distribution model extent is very small (e.g., less than 5 cells). Utilize the “All ecoregions” layer and the ERS ex situ legend to help locate these areas.

```{r echo=FALSE, message=FALSE, warning=FALSE}

start with a map of the distribution and ecoregions
if(dataNoModel == FALSE){
ex1 <- leaflet(width = "100%")|>
addProviderTiles(providersEsri.WorldGrayCanvas)|>
# Outline of all ecoregions
addPolygons(
data = allEcos,
group = "All ecoregions",
weight = 2,
stroke = TRUE,
fill = TRUE,
fillOpacity = 0.2,
popup = allEcosECO_NAME,
color = "#60616050"
) |>
leaflet::addRasterImage(x = raster(threshold),
colors = thresColor, 
group = "Buffered Area",
project = FALSE)
} else {
ex1 <- leaflet(width = "100%") |>
addProviderTiles(providersEsri.WorldGrayCanvas)
}

Add the GRSex maps
if(dataNoModel == FALSE){
if(exTablesTotal G with coordinates == 0){
grs_html <- HTML(
"
GRS Score: 0


No Map Elements
"
)

exG <- ex1|>
leaflet::addLegend(position = "bottomright",
        color = c( "#cfcccc","#45B320"),
        labels = c("Native Area", "Buffered Area"),
        title = "Modeled Habitat",
        group = "Buffered Area")|>
          addControl(
html = grs_html,
position = "bottomleft"  
)

}else{ # G points with valid lat lon are present
# generate the GRSex map
exG <- ex1 |>
#GRSex raster
addRasterImage(
x = raster(grsexMap),
colors = grsexColor,
group = "GRS ex situ",
project = FALSE)   |>
#G points
addCircleMarkers(
data = occData[occData$type == "G",],
color = ~ color,
opacity = 1,
radius = 1,
group = "GRS ex situ",
stroke = 1,
popup = ~ popup,
clusterOptions = markerClusterOptions()
) |>
#add legend
leaflet::addLegend(
position = "bottomright",
color = exPal,
title = "Ex Situ geographical representativeness",
labels = c("Native country-ecoregion area", "Buffered Area", "Areas < 50 km from an ex situ collection"),
group = "GRS ex situ"
)
}
}

Add the ERSex maps
if(data$NoModel == FALSE){

generate the ERSex map
if(datafcsexERS == 100 || length(ecoReg$ECO_ID_U) == 0){
ers_html <- HTML(
"
ERS Score: 100


No Map Elements
"
)

exE <- exG |>
  addControl(
html = ers_html,
position = "bottomleft")
} else {
  exE <- exG |>
  # Ecoregion Raster Layers
  addRasterImage(raster(ecoRast),
               group = "ERS ex situ",
               colors = ecoColor,
               project = FALSE) |>
  leaflet::addLegend(colors = ecoReg$color,
            group = "ERS ex situ",
            labels = ecoReg$ECO_NAME) |>
  # Outline of missing ecoregions
  addPolygons(
    data = ecoReg,
    group = "Missing ecoregions",
    color = "#f0a01f",  # This sets the outline color
    weight = 0.5,         # Outline width
    opacity = 0.5,        # Outline opacity
    fillOpacity = 0,    # Set to 0 for no fill
    popup = ~ECO_NAME)
}
}

add controls for the layout
ex3 <- exE |>
addLayersControl(
position = "topleft",
options = list(collapsed = FALSE,
autoZIndex = FALSE),
overlayGroups = c("Buffered Area",
"GRS ex situ",
"ERS ex situ",
"Missing ecoregions",
"All ecoregions"
)
) |>
hideGroup(c(
"ERS ex situ"
))

print the map
ex3

```

Ex situ conservation ecoregions table
The table below shows all the ecoregions overlapping with the distribution model of the taxon. The column ERS ex situ gap ecoregion denotes (true) if an ecoregion overlaps with the potential distribution model but is farther than 50 km from any ex situ conservation occurrence, and is thus considered a gap in ex situ conservation.

```{r echo=FALSE, message=FALSE, warning=FALSE}

All ecoregions considered
identify if they
have modeled habitat
are within a 50km buffer of a germplasm collection
if(dataNoModel == FALSE){
eco2 <- datanaturalArea |>
st_drop_geometry() |>
dplyr::mutate(
'ERSex Gap Ecoregion' = case_when(ECO_ID_U %in% missingEcos ~ TRUE,
TRUE ~ FALSE)
) |>
dplyr::select(
ERS ex situ gap ecoregion = 'ERSex Gap Ecoregion',
Unique ID = "ECO_ID_U",
Ecoregion ame = "ECO_NAME"  ) |>
arrange(desc(ERS ex situ gap ecoregion))
#visualize
DT::datatable(eco2,
rownames = FALSE,
class = "compact")

}else{
eco2 <- data$naturalArea |>
st_drop_geometry() |>
dplyr::mutate(
'ERSex Gap Ecoregion' =  TRUE) |>
dplyr::select(
ERS ex situ gap ecoregion = 'ERSex Gap Ecoregion',
Unique ID = "ECO_ID_U",
Ecoregion name = "ECO_NAME"  ) |>
arrange(desc(ERS ex situ gap ecoregion))
#visualize
DT::datatable(eco2,
rownames = FALSE,
class = "compact")
}

```

In situ conservation gap analysis
The table below shows the in situ conservation gap analysis summary. The Sampling Representativeness Score in situ (SRS in situ) calculates the proportion of all occurrences of a taxon within its predicted distribution that fall within a protected area. The Geographical Representativeness Score in situ (GRS in situ) compares the area (in km^2^) of the potential distribution model located within protected areas versus the total area of the model. The Ecological Representativeness Score in situ (ERS in situ) calculates the proportion of ecoregions encompassed within the range of the taxon located inside protected areas to the ecoregions encompassed within the total area of the potential distribution model. A Final Conservation Score for in situ (FCS in situ) was derived by calculating the average of the three in situ conservation metrics. All of the conservation metrics are on a scale from 0-100, with 0 = poor state of conservation and 100 = comprehensive conservation. The FCS in situ is used to categorize taxa, with urgent priority (UP) for further conservation action assigned when FCS in situ < 25, high priority (HP) where 25 ≤ FCS in situ < 50, medium priority (MP) where 50 ≤ FCS in situ < 75, and low priority (LP) for taxa whose FCS in situ ≥75.

```{r echo=FALSE, message=FALSE, warning=FALSE}

inCounts <- datacountsData∣>dplyr::mutate(across(c( 
′
 species 
′
 ),str 
r
​
 eplace 
a
​
 ll,pattern=" 
"
​
 ,replacement=""))∣>dplyr::mutate("Totaloccurrencesinprotectedareas"=round(datacountsDatatotalUseful∗(datafcsinSRS/100)))∣>dplyr::select("Taxon"=species,"Totaloccurrences"=totalUseful,"Totaloccurrencesinprotectedareas")if(dataNoModel == FALSE){
inScores <- data$fcsin |>
dplyr::mutate("SRS in situ" = sigfig(SRS),
"GRS in situ" = sigfig(GRS),
"ERS in situ" = sigfig(ERS),
"FCS in situ" = sigfig(FCS),
"Total occurrences in modeled area" = totalObsinThres,
"Total occurrences in protected areas" = totalObsinPro)|>
dplyr::select("SRS in situ","GRS in situ","ERS in situ","FCS in situ")|>
dplyr::mutate(
"FCS in situ priority category" = case_when(
FCS in situ >= 75 ~ "LP",
FCS in situ >=50 & FCS in situ < 75 ~ "MP",
FCS in situ >=25 & FCS in situ < 50 ~ "HP",
FCS in situ < 25 ~ "UP"
)
)

combine the data
inTables <- bind_cols(inCounts,inScores)
#visualize
DT::datatable(inTables,
rownames = FALSE,
class = "compact",
options = list(dom = 't'))
}else{
inScores <- datafcsin∣>dplyr::mutate("SRSinsitu"=sigfig(SRS),"GRSinsitu"=sigfig(GRS),"ERSinsitu"=sigfig(ERS),"FCSinsitu"=sigfig(FCS),"Totaloccurrencesinmodeledarea"=NA,"Totaloccurrencesinprotectedareas"=inCountsTotal occurrences in protected area)|>
dplyr::select("SRS in situ","GRS in situ","ERS in situ","FCS in situ")|>
dplyr::mutate(
"FCS in situ priority category" = case_when(
FCS in situ >= 75 ~ "LP",
FCS in situ >=50 & FCS in situ < 75 ~ "MP",
FCS in situ >=25 & FCS in situ < 50 ~ "HP",
FCS in situ < 25 ~ "UP"
)
)

combine the data
inTables <- bind_cols(inCounts,inScores)
#visualize
DT::datatable(inTables,
rownames = FALSE,
class = "compact",
options = list(dom = 't'))
}

```

Map of in situ conservation gap analysis measures
The SRS in situ map feature shows occurrences within the potential distribution that are outside of protected areas. Values are summarized by count per cell. The GRS in situ map feature shows the potential distribution, with distribution occurring within existing protected areas in purple and that outside of protected areas in green. The ERS in situ map features shows ecoregions within the potential distribution model that have no protected areas. There are many cases where the ecoregion area within the potential distribution model extent is very small (e.g., less than 5 cells). Utilize the “All ecoregions” layer and the ERS in situ legend to help locate these areas.

```{r echo=FALSE, message=FALSE, warning=FALSE}

generate the base map
if(dataNoModel == FALSE){
in1 <- leaflet(width = "100%")|>
addProviderTiles(providersEsri.WorldGrayCanvas)|>
# Outline of all ecoregions
addPolygons(
data = allEcos,
group = "All ecoregions",
weight = 2,
stroke = TRUE,
fill = TRUE,
fillOpacity = 0.2,
popup = allEcosECO_NAME,
color = "#60616050"
) |>
leaflet::addRasterImage(x = raster(threshold),
colors = thresColor, 
group = "Buffered Area",
project = FALSE)
} else {
in1 <- leaflet(width = "100%") |>
addProviderTiles(providersEsri.WorldGrayCanvas)
}

srsin map
if(data$NoModel == FALSE){
uniqueSRSVals <- sort(unique(values(srs1)))

condition for no values
if(length(uniqueSRSVals) == 1){
srs_html <- HTML(
"
SRS Score: 0


No Map Elements
"
)

inS <- in1|>
  addControl(
html = srs_html,
position = "bottomleft")
}else{

inS <- in1 |>

SRSin
addRasterImage(
x = raster(srs1),
colors = srsPal,
group = "SRS in situ",
project = FALSE)|>

protected Areas
addRasterImage(
x = raster(protectedArea),
colors = "#7570b3",
group = "SRS in situ",
project = FALSE )|>
leaflet::addLegend(colors = srsPal1,
group = "SRS in situ",
title = "SRS in situ",
labels = c(min(uniqueSRSVals),"", "","","",max(uniqueSRSVals)))|>
leaflet::addLegend(colors = "#7570b3",
group = "SRS in situ",
# title = "Protected Areas",
labels = c("Protected Areas"))
}
}

GRS map
if(data$NoModel == FALSE){
uniqueSRSVals <- sort(unique(values(srs1)))

condition for no values
if(datafcsinGRS == 0){
grs_html <- HTML(
"
GRS Score: 0


No Map Elements
"
)

inG <- inS|>
  addControl(
html = grs_html,
position = "bottomleft")
}else{

##GRSin map
inG <- inS |>
addRasterImage(
x = raster(grs1),
colors = grsinColor,
group = "GRS in situ",
project = FALSE
)|>
leaflet::addLegend(colors = exPal,
group = "GRS in situ",
labels = c("Background area", "Buffered Area", "Buffered Area within a protected area"))
}
}

ERS map
if(data$NoModel == FALSE){

condition for no values
if(datafcsinERS == 100 || class(ersinMap) == "character"){
ers_html <- HTML(
"
ERS Score: 100


No Map Elements
"
)

inE <- inG|>
  addControl(
html = ers_html,
position = "bottomleft")
}else{
missEcoIn <- allEcos[allEcosECO 
I
​
 D 
U
​
 ECO_ID_U, ]
##ERSin map
inE <- inG |>
addRasterImage(x = raster(ersinMap),
colors = ecoInColor,
group = "ERS in situ",
project = FALSE
)|>
leaflet::addLegend(colors = ers2color,group="ERSinsitu",labels=ers2ECO_NAME)|>
# Outline of all ecoregions
addPolygons(
data = missEcoIn,
group = "Missing ecoregions",
color = "#f0a01f",  # This sets the outline color
weight = 0.5,         # Outline width
opacity = 0.5,        # Outline opacity
fillOpacity = 0,    # Set to 0 for no fill
popup = ~ECO_NAME)
}
}

add control groups
if(data$NoModel == FALSE){
inMap <- inE |>
addLayersControl(
position = "topleft",
options = list(collapsed = FALSE,
autoZIndex = FALSE),
overlayGroups = c(
"Buffered Area",
"SRS in situ",
"GRS in situ",
"ERS in situ",
"Missing ecoregions",
"All ecoregions"))|>
hideGroup(group = c( "SRS in situ", "ERS in situ","All ecoregions"))

inMap
}else{
"No model was generate for this species"
}
```


In situ conservation ecoregions table
The table below shows all the ecoregions overlapping with the distribution model of the taxon. The column ERS in situ gap ecoregion denotes (true) if an ecoregion overlaps with the potential distribution model but none of that area is within a protected area, and is thus considered a gap in in situ conservation.

```{r echo=FALSE, message=FALSE, warning=FALSE}

All ecoregions considered
identify if they
have modeled habitat
are within a 50km buffer of a germplasm collection
if(dataNoModel == FALSE){
eco3 <- datanaturalArea |>
st_drop_geometry() |>
dplyr::mutate(
'ERS in situ gap ecoregion' = case_when(ECO_ID_U %in% ers2$ECO_ID_U ~ TRUE,
TRUE ~ FALSE)
) |>
dplyr::select(
'ERS in situ gap ecoregion',
Unique ID = "ECO_ID_U",
Ecoregion name = "ECO_NAME",
) |>
arrange(desc(ERS in situ gap ecoregion))
#visualize
DT::datatable(eco3,
rownames = FALSE,
class = "compact")
}else{
"No model was generated for this species"
}

```

Combined conservation gap analysis
The table below displays all conservation gap analysis scores, providing a comparison of the ex situ and in situ conservation scores.
```{r echo=FALSE, message=FALSE, warning=FALSE,fig.align='right',out.width="100%"}
exTable <- datafcsex∣>dplyr::mutate(across(c( 
′
 ID 
′
 ),str 
r
​
 eplace 
a
​
 ll,pattern=" 
"
​
 ,replacement=""))names(exTable)<−c("Taxon","ExsituSamplingRepresentativenessScore","ExsituGeographicRepresentativenessScore","ExsituEcologicalRepresentativenessScore","ExsituFinalConservationScore","ExsituConservationPriority")inTable<−datafcsin |>
dplyr::select(- ID)
names(inTable) <- c("In situ Sampling Representativeness Score",
"In situ Geographic Representativeness Score",
"In situ Ecological Representativeness Score",
"In situ Final Conservation Score",
"In situ Conservation Priority")

finalTable <- bind_cols(exTable, inTable)|>

round values in specific column
dplyr::mutate(across(c(2:5,7:10), round, digits = 2))

generate the combine conservation score
meanVal <- mean(c(finalTable$Ex situ Final Conservation Score, finalTable$In situ Final Conservation Score))

finalTable <- finalTable |>
dplyr::mutate(
Final Conservation Score Mean = round(meanVal, digits = 2),
Combined Conservation Priority = case_when(
meanVal <= 25 ~ "UP",
meanVal > 25 & meanVal <= 50 ~ "HP",
meanVal > 50 & meanVal <= 75 ~ "MP",
meanVal <= 75 ~ "UP",
)
)
#visualize
DT::datatable(finalTable,
rownames = FALSE,
class = "compact",
options = list(dom = 't'))

```


Conservation scores
The figure below displays all conservation gap analysis scores, providing a comparison of the ex situ and in situ conservation scores.

```{r echo=FALSE, message=FALSE, warning=FALSE,fig.align='right',out.width="100%"}


# Data preparation
d2 <- rbind(datafcsex, datafcsin) |>
  mutate(Category = c("ex situ", "in situ")) |>
  dplyr::select(
    ID,
    "SRS" = SRS,
    "GRS" = GRS,
    "ERS" = ERS,
    "FCS" = FCS,
    "Score Class" = FCS_Score,
    Category
  )

d3 <- d2 |>
  pivot_longer(cols = c("SRS", "GRS", "ERS", "FCS"),
               names_to = "Metric")

# Define the combined score for the species
combined_mean <- as.numeric(mean(c(d3value[4], d3value[8])))

d3[9, ] <- list(
  d3$ID[1],
  NA,
  "Combined conservation score",
  "Combined Final


Conservation Score",
combined_mean
)

# Define the combined score category
if (d$3value[9] <= 25) {
  d3$Score Class[9] <- "UP"
} else if (d3value[9] <= 50) {
  d3$Score Class[9] <- "HP"
} else if (d3value[9] <= 75) {
  d3$Score Class[9] <- "MP"
} else {
  d3$Score Class[9] <- "LP"
}

# Update factor levels to match the new string with the line break
d3M$etric < −factor(d3$Metric,
                   levels = c("SRS", "GRS", "ERS", "FCS", "Combined Final


Conservation Score"))

Plotting
plot_ly(d3,
        x = ~ Metric,
        y = ~ value,
        color = ~ Category) |>
  layout(
    xaxis = list(
      title = "",
      tickangle = 0,
      
      # --- Add these two lines to force the order ---
      categoryorder = "array",
      categoryarray = c(
        "SRS",
        "GRS",
        "ERS",
        "FCS",
        "Combined Final<br>Conservation Score"
      ),
      # ----------------------------------------------
      
      tickmode = "array",
      tickvals = c(
        "SRS",
        "GRS",
        "ERS",
        "FCS",
        "Combined Final<br>Conservation Score"
      ),
      ticktext = c(
        "              SRS",
        "              GRS",
        "              ERS",
        "              FCS",
        "Combined Final<br>Conservation Score"
      )
    ),
    yaxis = list(
      title = "",
      range = c(0, 100) # Locks the y-axis to always show 0 to 100
    ),
    margin = list(b = 60),
    barmode = 'group',
    bargroupgap = 0,
    # Pushes the bars together
    hovermode = 'compare'
  )
```

Note: Score values of zero will not appear on the chart.

Definitions of conservation gap analysis scores
Sampling Representativeness Score (SRS)
Ex situ: The Sampling Representativeness Score ex situ (SRS ex situ) calculates the ratio of germplasm accessions (G) available in ex situ repositories to reference/voucher/herbarium (H) records for each taxon, making use of all compiled records irrespective of whether they include coordinates.

In situ: The Sampling Representativeness Score in situ (SRS in situ) calculates the proportion of all occurrences of a taxon within its native range that fall within a protected area.

Geographic Representativeness Score (GRS)
Ex situ: The Geographic Representativeness Score ex situ (GRS ex situ) uses 50-km-radius buffers created around each G collection coordinate point to estimate geographic areas already well collected within the potential distribution models of each taxon and then calculates the proportion of the potential distribution model covered by these buffers.

In situ: The Geographic Representativeness Score in situ (GRS in situ) compares the area (in km^2^) of the potential distribution model located within protected areas versus the total area of the model.

Ecological Representativeness Score (ERS)
Ex situ: The Ecological Representativeness Score ex situ (ERS ex situ) calculates the proportion of terrestrial ecoregions represented within the G buffered areas out of the total number of ecoregions occupied by the potential distribution model.

In situ: The Ecological Representativeness Score in situ (ERS in situ) calculates the proportion of ecoregions encompassed within the range of the taxon located inside protected areas to the ecoregions encompassed within the total area of the potential distribution model.

Final Conservation Score (FCS)
Ex situ: The Final Conservation Score ex situ (FCS ex situ) was derived by calculating the average of the three ex situ conservation metrics.

In situ: The Final Conservation Score in situ (FCS in situ) was derived by calculating the average of the three in situ conservation metrics.

FCSc-mean: The Combined Final Conservation Score (FCSc-mean) was calculated for each taxon by averaging its final FCS ex situ and FCS in situ scores.

Taxa were further categorized with regard to the two conservation strategies as well as in combination, with Urgent Priority (UP) for further conservation action assigned when FCS < 25, High Priority (HP) assigned when 25 ≤ FCS < 50, Medium Priority (MP) when 50 ≤ FCS < 75, and Low Priority (LP) when FCS ≥75.

Definitions of occurrence data categories
  
Occurrences: The total number of observations and ex situ collections records for the taxon that were evaluated for use within the study.

Occurrences with lat/long: The total number of observations that had a valid latitude and longitude pair and were used for spatial analysis. Non-valid latitude longitude pairs included but were not limited to points not on land masses, and records with only one of the two values present.

Germplasm Records (G): Occurrences in which a living sample (typically a live plant or seed) is present in an ex situ conservation system (i.e., botanical garden, seed bank, genebank, etc.).

Reference/voucher Records (H): Occurrences that have a supporting herbarium or other reference record.

Unique Data Sources: The number of unique database sources from which occurrences of the taxon were gathered.